---
title: "gbm model"
format: html 
editor: visual
Notes: This script is intended to do initialise the building of the gbm ml model using the input and ouput data with multiprocessing.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

## Load the libraries

## Load the libraries

```{r}
library(dplyr)
library(readr)
library(mltools)
library(knitr)
library(factoextra)
library(data.table)
library(psych)
library(limma)
library(edgeR)
library(tidyverse)
library(caret)
```

Re-usable functions

```{r}
extract_element <- function(strings, index) {
  # Split each string by "." and extract the third element
  element_list <- sapply(strsplit(strings, "\\."), function(x) x[index])
  return(element_list)
}

untransform_exp <- function(x) {#Function to convert the log transformed counts back into original counts
  return(ceiling((2^x)-1))
}

transform_exp <- function(x) {#Function to convert the log transformed counts back into original counts
  return(log2(x + 1))
}

untransform_tpm <- function(x) {#Function to convert the log transformed counts back into original counts
  return(ceiling((2^x)-0.001))
}

transform_tpm <- function(x) {#Function to convert the log transformed counts back into original counts
  return(log2(x + 0.001))
}
```

Import data

```{r}
# SOI genes
# soi <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
# soi_genes <-soi[,2]
# rm(soi)

# Metadata
# Sample meta
# tss_meta <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")
# abbrv_meta <- read.csv("Data/Other/TCGA_meta/bcrBatchCode.tsv", sep = "\t")

# Gene Metadata
gene_ids <- read.delim("Data/Other/TCGA_meta/TCGA_PanCan_TPM_Gene_Annotations.txt")

# Predictor variables
# Expected Counts
# ori_exp <- read.csv("Data/RNA_Data/TCGA_Norm/tcga_gene_expected_count.csv")
order_exp <- ori_exp[,order(colnames(ori_exp))]

# Convert the Gene Ids into names
order_exp <- right_join(gene_ids %>% 
                    dplyr::select(c("id", "gene")) %>% 
                    sapply(trimws) %>% 
                    as.data.frame(),
                  order_exp,
                  by = c("id" = "sample")) %>% 
  dplyr::select(-"id") %>% 
  rename(Gene = "gene")

# TPM counts
# ori_tpm <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_Full.csv")
order_tpm <- ori_tpm[,order(colnames(ori_tpm))] %>% 
  dplyr::select(-"id")

# Number of random rows and columns to select
n_extra_genes <- 20000
n_samples <- 3000

tpm_not_soi <- order_tpm %>%
  filter(!Gene %in% soi_genes) %>%
  sample_n(size = n_extra_genes, replace = FALSE)

order_tpm_soi <- order_tpm[order_tpm$Gene %in% soi_genes, ]

exp_not_soi <- order_exp %>%
  filter(!Gene %in% soi_genes) %>%
  sample_n(size = n_extra_genes, replace = FALSE)

order_exp_soi <- order_exp[order_exp$Gene %in% soi_genes, ]

order_tpm <- rbind(tpm_not_soi, order_tpm_soi)[,0:n_samples]
order_exp <- rbind(exp_not_soi, order_exp_soi)[,0:n_samples]

# Response data

# HRD scores
ori_hrd <- read_tsv("Data/CIN_Features/TCGA.HRD_withSampleID.txt")

# Arm level aneuploidies
ori_arm_cnv <- read_tsv("Data/CIN_Features/CNV_Data/PANCAN_ArmCallsAndAneuploidyScore_092817.txt")


rm(order_tpm_soi)
rm(tpm_not_soi)
rm(exp_not_soi)
rm(order_exp_soi)
rm(n_samples)
rm(n_extra_genes)

```

```{r}
# Combine the metadata
# meta <- left_join(tss_meta %>% 
                   #   dplyr::select(c("TSS.Code", "Study.Name")) %>% 
                   #   distinct() %>% 
                   #   sapply(trimws) %>% 
                   #   as.data.frame(),
                   # abbrv_meta %>%
                   #   dplyr::select(c("Study.Abbreviation", "Study.Name")) %>% 
                   #   distinct()%>% 
                   #   sapply(trimws) %>% 
                   #   as.data.frame(), 
                   # by = "Study.Name")
# rm(tss_meta)
# rm(abbrv_meta)
```

1.  Handle the configuration of the RNA seq data

```{r}
trans_exp <- order_exp # Log transformed Expected Counts
rownames(trans_exp) <- NULL
trans_tpm <- order_tpm # Log transformed TPM Counts
rownames(trans_tpm) <- NULL

count_exp <- trans_exp %>%
  mutate_at(vars(-1), untransform_exp)
count_tpm <- trans_tpm %>%
  mutate_at(vars(-1), untransform_tpm)

# Remove the non-cancerous sample types from the set
codes_to_use <- c("01","02","03","04","05","08","09")

exp_samples_to_use <- count_exp %>%  dplyr::select(c("Gene", ends_with(codes_to_use)))
tpm_samples_to_use <- count_tpm %>%  dplyr::select(c("Gene", ends_with(codes_to_use)))

counts_exp <- tpm_samples_to_use %>%
  mutate(Gene = trimws(Gene))

# Convert data to a data.table for faster processing during the grouping of the duplicate genes
setDT(counts_exp)  
setDT(tpm_samples_to_use)  

# Combine duplicate genes together using the median of the expression
grouped_exp <- counts_exp[, lapply(.SD, function(x) if (length(x) > 1) median(x, na.rm = TRUE) else x), by = Gene, .SDcols = -"Gene"]
grouped_tpm <- tpm_samples_to_use[, lapply(.SD, function(x) if (length(x) > 1) median(x, na.rm = TRUE) else x), by = Gene, .SDcols = -"Gene"]

# Reset the row names to the Gene names
groups_exp <- as.data.frame(grouped_exp)
groups_tpm <- as.data.frame(grouped_tpm)
```

```{r}
exp_data <- distinct(groups_exp) %>% 
  column_to_rownames(var = "Gene")
tpm_data <- distinct(groups_tpm) %>% 
  column_to_rownames(var = "Gene")

# Remove samples (cols) with 0 values throughout
cols_zeros_tpm <- which(apply(tpm_data, 2, function(x) all(x == 0)))
cols_zeros_exp <- which(apply(exp_data, 2, function(x) all(x == 0)))

if (length(cols_zeros_tpm) > 0) {
  data_complete_tpm <- tpm_data[, -cols_zeros_tpm]
} else {
  data_complete_tpm <- tpm_data
}

if (length(cols_zeros_exp) > 0) {
  data_complete_exp <- exp_data[, -cols_zeros_exp]
} else {
  data_complete_exp <- exp_data
}

# Identify genes with 0 expression across all samples
zero_genes_exp <- rowSums(data_complete_exp == 0) == ncol(data_complete_exp)

# Remove genes with 0 expression across all samples
filt_dt_exp <- data_complete_exp[!zero_genes_exp, ]
rm(zero_genes_exp)

# Turn into a DGE object
matrix_exp <- filt_dt_exp %>% as.matrix()
d_exp <- DGEList(matrix_exp)

# Calculate the normalisation factor 
Normfact_exp <- calcNormFactors(d_exp, method = "TMM")
Normfactors <- as.data.frame(Normfact_exp$samples) %>% select("norm.factors")

# match the column names from the normalisation factor df with that of the other dfs in order to divide the count values by the library size factors for the correct samples 
matching_cols_tpm <- colnames(tpm_data)[match(rownames(Normfactors), colnames(tpm_data))] 

rm(d_exp)
rm(Normfact_exp)
rm(filt_dt_exp)
rm(matrix_exp)
rm(order_exp)
rm(order_tpm)
rm(trans_exp)
rm(trans_tpm)
rm(gene_ids)
rm(exp_samples_to_use)
rm(codes_to_use)
rm(counts_exp)
rm(tpm_samples_to_use)
rm(grouped_exp)
rm(grouped_tpm)
rm(groups_exp)
rm(groups_tpm)
```

```{r}
matching_cols_tpm <- matching_cols_tpm[!is.na(matching_cols_tpm)]

scld_cnts_tpm <- round(sweep(tpm_data[, matching_cols_tpm], 2, Normfactors$norm.factors, "/"),2)

matching_cols_exp <- colnames(exp_data)[match(rownames(Normfactors), colnames(exp_data))]
matching_cols_exp <- matching_cols_exp[!is.na(matching_cols_exp)]

scld_cnts_exp <- round(sweep(exp_data[, matching_cols_exp], 2, Normfactors$norm.factors, "/"),2)

# Expected Counts
exp_set <- data_complete_exp %>% # Raw expected counts
  filter(rownames(.) %in% soi_genes) %>%
  t() %>% 
  as.data.frame()

scld_exp_set <- scld_cnts_exp %>% # Library size normalised Expected counts
  filter(rownames(.) %in% soi_genes) %>% 
  t() %>% 
  as.data.frame()

log_exp <- exp_set %>% # Log transformed expected counts
  mutate_at(vars(everything()), transform_exp) 

log_scld_exp <- scld_exp_set %>% # Log transformed Library size normalised expected counts
  mutate_at(vars(everything()), transform_exp) 

# Transcript per million (TPM)
tpm_set <- data_complete_tpm %>% # Raw TPM counts 
  filter(rownames(.) %in% soi_genes) %>% 
  t() %>% 
  as.data.frame()

scld_tpm_set <- scld_cnts_tpm %>% # Library size normalised TPM counts
  filter(rownames(.) %in% soi_genes) %>% 
  t() %>% 
  as.data.frame()

log_tpm <- tpm_set %>% # Log transformed TPM counts
  mutate_at(vars(everything()), transform_tpm) 

log_scld_tpm <- scld_tpm_set %>% # Log transformed Library size normalised TPM counts
  mutate_at(vars(everything()), transform_tpm) 

rm(scld_cnts_tpm)
rm(data_complete_tpm)
rm(scld_cnts_exp)
rm(data_complete_exp)
rm(exp_data)
rm(tpm_data)
```

2.  Handle the configuration of the CIN-based features

```{r}

t_hrd <- as.data.frame(t(ori_hrd)) 
first_hrd <- t_hrd
colnames(first_hrd) <- t_hrd[1, ]
hrd <- as.data.frame(first_hrd[-1,]) %>% 
  mutate_all(as.numeric) %>% 
  rename(loh_hrd = "hrd-loh")

# ARM-LEVEL ANEUPLOIDIES
# Replace the NAs with 0
cvn_arm <- ori_arm_cnv %>% replace(is.na(.), 0)

# Re-organise the columns 
cnvs_arm <- cvn_arm %>%
  column_to_rownames("Sample") %>% 
  dplyr::select(-"Type") %>% 
  mutate_all(as.numeric)

full_cin <- merge(
  hrd,
  cnvs_arm,
  by = "row.names"
) %>% 
  mutate(Row.names = str_replace_all(Row.names, "-", ".")) %>%
  column_to_rownames("Row.names")

rm(t_hrd)
rm(first_hrd)
rm(ori_arm_cnv)
rm(cvn_arm)
rm(hrd)
rm(cnvs_arm)
```

```{r}
n.cores <- parallel::detectCores() - 1
my.cluster <- parallel::makeCluster(
  n.cores, 
  type = "FORK"
  )

print(my.cluster)
```

4.  Complete merge of all the predictor and the "categorical" aneuploidy vars

```{r}
rna_list <- list(
  transcripts_per_million = tpm_set, # Seems to be the best performing
  scalled_transcripts_per_million = scld_tpm_set, # not too useful (scalled)
  log_scalled_transcripts_per_million = log_scld_tpm,
  log_transcripts_per_million = log_tpm,
  expected_counts = exp_set,
  scalled_expected_counts = scld_exp_set,
  log_expected_counts = log_exp,
  log_scalled_expected_counts = log_scld_exp 
)

# aneu_feature_list <- colnames(full_cin[1,6:length(full_cin)])
aneu_feature_list <- colnames(full_cin[1,20:24])
reg_feature_list <- colnames(full_cin[1,1:5])


```

```{r}
set.seed(123)
for (feature in aneu_feature_list){
  cat(paste0("\n", feature, ":"))
  
  aneu_cat_metrics_df <- data.frame(
    Feature = character(),
    RNA_Set = character(),
    Depth = numeric(),
    Learning_Rate = numeric(),
    Gamma = numeric(),
    Logloss = numeric()
  ) 
  
  for (i in 1:length(rna_list)) {
    
    rna <- rna_list[[i]]
    name <- names(rna_list)[i]
    
    cat(paste0("\n\t", name, ":"))
    
    full_df <- merge(rna, full_cin, by = "row.names")
    
    y <- as.integer(full_df[[feature]])
    X <- full_df %>% select(-append("Row.names",colnames(full_cin)))
    
    y[y == -1] <- 0
    y[y == 1] <- 2
    y[y == 0] <- 1
    
    xgb_data <- xgb.DMatrix(data = as.matrix(X), label = y)
    
    grid <- expand.grid(
      max_depth = seq(1, 15,4),
      gamma = seq(0, 2, 2.3),
      eta = seq(0.01, 0.1, 1)
    )
    
    for (j in 1:nrow(grid)) {
      cat("\n Depth ",grid$max_depth[j], 
          "\n Eta ", grid$eta[j], 
          "\n Gamma ", grid$gamma[j], "\n")
      m_xgb_untuned <- xgb.cv(
        data = xgb_data,
        nrounds = 1000,
        objective = "multi:softmax",
        early_stopping_rounds = 50,
        nfold = 3,
        max_depth = grid$max_depth[j],
        eta = grid$eta[j],
        num_class = 3,
        verbose = 0
      )
      
      
      best_loss <- m_xgb_untuned$evaluation_log$test_mlogloss_mean[m_xgb_untuned$best_iteration]
      
      aneu_cat_metrics_df <- rbind(aneu_cat_metrics_df,
                                   data.frame(
                                     Feature = feature,
                                     RNA_Set = name,
                                     Depth = grid$max_depth[j],
                                     Learning_Rate = grid$eta[j],
                                     Gamma = grid$gamma[j],
                                     Logloss = best_loss
                                   )
      )
      
    }
  }
  write.csv(aneu_cat_metrics_df, paste0("Data/aneu_cat_xgb_metrics_params_d",feature,"_",Sys.Date(), ".csv"))
}

```

5.  Complete merge, regression model and results for all the arm level aneuploidies as "continuous" vars instead of categorical

```{r}


for (feature in reg_feature_list){
  cat(paste0("\n", feature, ":"))
  
  aneu_reg_metrics_df <- data.frame(
    Feature = character(),
    RNA_Set = character(),
    Depth = numeric(),
    Learning_Rate = numeric(),
    Gamma = numeric(),
    RMSE = numeric()
  )

  for (i in 1:length(rna_list)){
    
    rna <- rna_list[[i]]
    name <- names(rna_list)[i]
    
    cat(paste0("\n\t", name, ":"))
    
    full_df <- merge(rna, full_cin, by = "row.names")
  
    y <- as.integer(full_df[[feature]])
    X <- full_df %>% select(-append("Row.names",colnames(full_cin)))
    
    y[y == -1] <- 0
    y[y == 1] <- 2
    y[y == 0] <- 1
    
    xgb_data <- xgb.DMatrix(data = as.matrix(X), label = y)
    
    grid <- expand.grid(
          max_depth = seq(1, 15,4),
          gamma = seq(0, 2, 2.3),
          eta = seq(0.01, 0.1, 1)
    )
        
    for (j in 1:nrow(grid)) {
      cat("\n Depth ",grid$max_depth[j], 
          "\n Eta ", grid$eta[j], 
          "\n Gamma ", grid$gamma[j], "\n")
      set.seed(123)
      m_xgb_untuned <- xgb.cv(
        data = xgb_data,
        nrounds = 2000,
        objective = "reg:squarederror",
        eval_metric = "rmse",
        early_stopping_rounds = 50,
        nfold = 3,
        max_depth = grid$max_depth[j],
        eta = grid$eta[j],
        verbose = 0
      )
      
      best_rmse <- m_xgb_untuned$evaluation_log$test_rmse_mean[m_xgb_untuned$best_iteration]
      
      aneu_reg_metrics_df <- rbind(aneu_reg_metrics_df,
                                data.frame(
                                  Feature = feature,
                                  RNA_Set = name,
                                  Depth = grid$max_depth[j],
                                  Learning_Rate = grid$eta[j],
                                  Gamma = grid$gamma[j],
                                  RMSE = best_rmse
                                  )
                                )
      
        
      }
    }
  
  write.csv(aneu_reg_metrics_df, paste0("Data/aneu_reg_xgb_metrics_params_",feature,"_",Sys.Date(), ".csv"))
  
}

```
