---
title: "TCGA_explore"
output: html_document
date: "2024-02-19"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```
Load the libraries
```{r}
library(dplyr)
library(plyr)
library(knitr)
library(tidyverse)
``` 

Load other options
```{r}
getOption("max.print")
```


```{r}
sampletype <- read.delim("Data/Other/TCGA_meta/sampleType.tsv", sep = "\t", colClasses = c("character"))
tissuesource <- read.delim("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")
TCGAtpm <-read.delim("Data/RNA_Data/TCGA_TPM/tcga_RSEM_gene_tpm.txt", header = TRUE, sep = "\t", nrows = 4)
```



```{r}
samples <- TCGAtpm[1, 2:ncol(TCGAtpm)] %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "Samples") %>%
  mutate(type = strsplit(Samples, "\\.") %>% sapply("[[", 4), 
         partic = strsplit(Samples, "\\.") %>% sapply("[[", 3)) %>%
  inner_join(sampletype, by = c("type" = "Code"))
```


Check the total number of total Normal tissue types
```{r}
types <- count(samples$Definition, wt = NULL) 
kable(types)
```

Total of 727 total normal type solid tissue samples


Check how many of the 727 samples are duplicates from the same individual
```{r}
normalall <- samples[samples$Definition== "Solid Tissue Normal", ] %>% count("partic")
```

None of the normal type solid tissue samples are duplicated within this set




Check how many of the 727 have cancerous duplicates
```{r}

subset_df <- samples[duplicated(samples$partic) | duplicated(samples$partic, fromLast = TRUE),]
normaldupli <- count(subset_df$Definition)
kable(normaldupli)
```
681 of the 727 non-cancerous samples have cancerous duplicates 


Determine what the duplicates for the 681 normal samples are
```{r}
normalduplipart <- subset_df[subset_df$Definition == "Solid Tissue Normal", "partic"]
duplinormals <-subset_df[subset_df$partic %in%  normalduplipart,]
particduplinormalscount <- duplinormals %>% dplyr::count(partic, sort = TRUE)
kable(head(particduplinormalscount))
```





```{r}
defduplinormalscount <- duplinormals %>% dplyr::count(Definition, sort = TRUE)
kable(head(defduplinormalscount))
```
681 normal tissue type samples with corresponding (duplicate) tissue types
680 of the 681 have a primary solid tumor duplicate
4 of the 681 have a metastatic duplicate
3 of the 681 are triplicates with both a metastatic and a primary solid tumor duplicate


get the full list of samples whose participants have both a cancer sample and a control 
```{r}
metastatic_participants <- duplinormals[duplinormals$Definition == "Metastatic", "partic"]
usable_participants <-subset_df[subset_df$partic %in%  normalduplipart & !(subset_df$partic %in%  metastatic_participants),]

```

677 usable participants with both Normal Tissue and Cancer Tissue without metastatic 

