---
title: "CNV pre-analysis"
output: html_document
date: "2024-01-09"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```
Load the libraries
```{r}
library(dplyr)
library(broom)
library(knitr)
library(coin)
library(data.table)
library(ggpubr)
library(rstatix)
# library(ggplot2)
# library(paletteer)
# library(PairedData)
library(tidyverse)
``` 

Load other options
```{r}
getOption("max.print")
```


Load the CNV metrics dataset and remove the useless columns. 

The data shows the samples as rows and descriptive characteristics, the Tissue Source Site (TSS) and the Histology grouping as columns.

```{r}
random_rows <- sample(1:11060, 600)


first <- fread("Data/CNV_Data/Full_CNV_metrics1.csv")

used_first <- first[,c("Sample", "Mean_set", "Mean_cntrl","TSS","Hist_group", "TSS_Abbrvs")]
setnames(used_first, old = "Mean_cntrl", new = "Mean_cntrl1")
used_first[, Mean_diff1 := (Mean_set - Mean_cntrl1)]

num_datasets <- 412

# Loop over the datasets
for (i in 2:num_datasets) {
  
  file_name <- paste("Data/CNV_Data/Full_CNV_metrics", i, ".csv", sep = "")
  

  # Read the CSV file into a data.table
  current_dataset <- fread(file_name)
  current_dataset[, Mean_diff := (Mean_set - Mean_cntrl)]

  new_name1 = paste('Mean_cntrl', i, sep = "")
  new_name2 = paste('Mean_diff', i, sep = "")

  rename_map <- c("Mean_cntrl" = new_name1, "Mean_diff" = new_name2)

  # Rename columns in 'current_dataset' using the named vector
  setnames(current_dataset, old = names(rename_map), new = rename_map)

  # Select columns with their original names from 'current_dataset' and rename them
  selected_columns <- current_dataset[random_rows, c("Sample", new_name1, new_name2), with = FALSE]

  # Perform a merge based on Sample ID
  used_first <- merge(used_first, selected_columns, by = "Sample", all = FALSE, suffixes = c("", i))

}


``` 




QUESTION 1 


Visualise the distribution of the set of interest

```{r}
ggplot(used_first, aes(x = Mean_set)) +
  geom_histogram(aes(y = after_stat(density)),
                 colour = 1,
                 fill = "gray88",
                 bins = 50) +
  geom_density(lwd = 1,
               colour = 1) +
  labs(y = "Density", x = "CNV log2 ratio",
       title = "CNV Distribution of the Set-of-Interest") +
  theme(
    axis.title = element_text(size = 13, color = "black", face = "bold"),
    axis.text = element_text(color = "black", size = 10),
    panel.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.6, size = 15),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.major.x = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.minor.x = element_line(color = "black", linewidth = 0.1, linetype = 1),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 1),
    plot.title.position = "plot"
  )
```
Interpretation: The distribution of the CNV for the SoI is clearly Leptokurtic with a non-normal distribution




Visualise the distribution of the control sets 

```{r}
sample_control_sets <- used_first %>% 
  gather(key = "Control", value = "Mean_CNV", starts_with("Mean_cntrl39")) %>%
  mutate(Sample = as.factor(Sample), Control = as.factor(Control)) %>%
  select(Sample, Control, Mean_CNV)

ggplot(sample_control_sets, aes(x = Mean_CNV, group = Control)) +
  geom_histogram(aes(y = after_stat(density)),
                 colour = 1,
                 fill = "white",
                 bins = 50) +
  geom_density(lwd = 1,
               colour = 1) +
  facet_wrap(~Control) +
  labs(y = "Density", x = "CNV log2 ratio",
       title = "CNV Distribution of the control set") +
  theme(
    axis.title = element_text(size = 13, color = "black", face = "bold"),
    axis.text = element_text(color = "black", size = 10),
    panel.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.6, size = 15),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.major.x = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.minor.x = element_line(color = "black", linewidth = 0.1, linetype = 1),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 1),
    plot.title.position = "plot"
  )
```
Interpretation: The distribution of the CNV for the controls sets appears also Leptokurtic with a non-normal distribution. The fact that both the controls and the SoI share the same distribution is indicative that standard non-parametric tests should be able to be applied for comparisons. 

```{r}
sample_control_sets <- used_first %>% 
  gather(key = "Control", value = "Mean_CNV_dif", starts_with("Mean_diff39")) %>%
  mutate(Sample = as.factor(Sample), Control = as.factor(Control)) %>%
  select(Sample, Control, Mean_CNV_dif)

ggplot(sample_control_sets, aes(x = Mean_CNV_dif, group = Control)) +
  geom_histogram(aes(y = after_stat(density)),
                 colour = 1,
                 fill = "white",
                 bins = 50) +
  geom_density(lwd = 1,
               colour = 1) +
  facet_wrap(~Control) +
  labs(y = "Density", x = "CNV log2 ratio",
       title = "CNV Distribution of the control set") +
  theme(
    axis.title = element_text(size = 13, color = "black", face = "bold"),
    axis.text = element_text(color = "black", size = 10),
    panel.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.6, size = 15),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.major.x = element_line(color = "black", linewidth = 0.25, linetype = 1),
    panel.grid.minor.x = element_line(color = "black", linewidth = 0.1, linetype = 1),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 1),
    plot.title.position = "plot"
  )
```
Interpretation: The distribution of the differences in CNV values across different pairings appears to be somewhat symmetrical but with a positive bulge/skew. This could indicate that the gene-set of interest contains increased copy number gains compared to the control gene-sets.






