
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```
Load the libraries
```{r}
library(dplyr)
library(broom)
library(knitr)
library(coin)
library(data.table)
library(ggpubr)
library(rstatix)
# library(ggplot2)
# library(paletteer)
# library(PairedData)
library(tidyverse)
library(jmuOutlier)
library(paletteer)
``` 

Load other options
```{r}
# options(scipen = 1, digits = 20)
getOption("max.print")
```


Load the RNA metrics dataset and remove the useless columns. 

The data shows the samples as rows and descriptive characteristics, the Tissue Source Site (TSS) and the Histology grouping as columns.

```{r}
set.seed(123)
random_rows <- sample(1:11060, 1000)

first <- fread("Data/RNA_Data/Full_RNA_metrics1.csv")

used_first <- first[random_rows,c("Sample", "Mean_set", "Mean_cntrl","TSS","Hist_group", "TSS_Abbrvs")]
setnames(used_first, old = "Mean_cntrl", new = "Mean_cntrl1")
used_first[, Mean_diff1 := (Mean_set - Mean_cntrl1)]

num_datasets <- 412

# Loop over the datasets
for (i in 2:num_datasets) {
  
  file_name <- paste("Data/RNA_Data/Full_RNA_metrics", i, ".csv", sep = "")
  

  # Read the CSV file into a data.table
  current_dataset <- fread(file_name)
  current_dataset[, Mean_diff := (Mean_set - Mean_cntrl)]

  new_name1 = paste('Mean_cntrl', i, sep = "")
  new_name2 = paste('Mean_diff', i, sep = "")

  rename_map <- c("Mean_cntrl" = new_name1, "Mean_diff" = new_name2)

  # Rename columns in 'current_dataset' using the named vector
  setnames(current_dataset, old = names(rename_map), new = rename_map)

  # Select columns with their original names from 'current_dataset' and rename them
  selected_columns <- current_dataset[random_rows, c("Sample", new_name1, new_name2), with = FALSE]

  # Perform a merge based on Sample ID
  used_first <- merge(used_first, selected_columns, by = "Sample", all = FALSE, suffixes = c("", i))

}

print(head(used_first))
``` 


Choose a random set of 5 control gene-sets that will be used in the graphs

```{r}
set.seed(123)
controls <- sample(1:num_datasets, 5)
```





Part I. Inter-genomic Expression analysis 


QUESTION 1 


A) Create violin plots of the differences of the average log2 expression to assess the type of paired test to perform on the the pairwise groups .

```{r}

dif_pairing_names <- unlist(lapply(controls, function(i) paste("Mean_diff", i, sep ="")))

violin_graph_long_data <- used_first %>% 
  pivot_longer(cols = dif_pairing_names,
               names_to = "Variable",
               values_to = "Value") 

ggplot(violin_graph_long_data, aes(y = Value, x = Variable)) +
  geom_violin(trim = FALSE, fill = "gray88")+
  geom_boxplot(varwidth = T, alpha  = 0.25, outlier.colour = "red")+
  labs(x = "Unique SoI / Control Pairings ", y = "Δ Log2 Expression",
       title = "Distribution of the Set-of-Interest to Control Gene-Set \n  Expression Difference Across Various Pairings") +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 13, color = "black", face = "bold"),
    axis.text = element_text(color = "black", size = 10),
    axis.text.x = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.text = element_blank(),
    plot.title = element_text(hjust = 0.6, size = 15),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    plot.title.position = "plot"
  )


``` 
Interpretation: Plots clearly show distinct pairings which violate a normal distribution. This follows that a non-parametric test of paired differences between the means can be performed for all pairings. This method is more stringent but more efficient than evaluating each distribution individually to decide on whether to apply a parametric or non-parametric test to the distributions. 
The data is however not symmetrical around the median and the paired Wilcoxon test has this as an assumption which will come into play for the statistical comparisons between the SoI and control gene-sets. Should not be an issue when comparing amongst each other though as the skew is in the same direction for all pairings. 
The plots also demonstrate that the variance of the different pairings (width) is quite similar despite the fact that the shape may deviate somewhat between pairings. 


B) Use a Friedman's test (non-parametric version of the RMA, which is a paired t-test for more than 2 groups) to test whether the different control groups differ significantly in their distributions.


```{r}
# Fried_test_data <- used_first %>%
#   gather(key = "Pairing", value = "Difference", starts_with("Mean_cntrl")) %>%
#   mutate(Sample = as.factor(Sample), Pairing = as.factor(Pairing)) %>% 
#   select(Sample, Pairing, Difference)
# 
# friedman_result <- coin::friedman_test(Difference ~ Pairing | Sample, data = Fried_test_data)
# print(friedman_result)
```
Result: The Friedman Test shows that at least one of the pairings demonstrates a significantly different distribution. However, performing pairwise post hoc tests between all pairs of groups may not be practical due to the combinatorial explosion of comparisons


C) Construct a contrast analysis to assess each of the pairings against the general trend

```{r}
# Paired_wilc_table <- used_first %>% 
#   dplyr::select(starts_with("Mean_diff")) %>%
#   mutate_all(as.numeric) 
#    
# Contrast_table <- data.frame(Set = character(0), P_Value = numeric(0), Adj_P_Value = numeric(0), Significant = character(0))
# 
# # Loop over each control gene-sets
# for (i in 1:num_datasets) {
#   
#   set = paste('Mean_diff', i, sep = "")
#   
#    Global_table <- Paired_wilc_table %>%
#     mutate(global = rowMeans(select(., -set), na.rm = TRUE)) %>%
#     select(set, global)
#    
#    wlcx_paired_res = wilcox.test(Global_table[,get(set)], Global_table$global, paired = TRUE)
#    Corrected_Pval <- wlcx_paired_res$p.value * num_datasets
#    Significant_value <- if(Corrected_Pval < 0.05) "TRUE" else "FALSE"
#    
#    Contrast_table <- rbind(Contrast_table, data.frame(Set = set, P_Value = wlcx_paired_res$p.value, Adj_P_Value = Corrected_Pval, Significant = Significant_value)) %>% arrange(desc(Adj_P_Value))
# }

# Contrast_agg_count_table <- Contrast_table %>% group_by(Significant) %>% summarise(count = n())
# print(Contrast_agg_count_table)
```
Result: The Contrast analysis shows that the expression pattern of the control gene-sets seem to mostly differ significantly compared to their average with only 25/412 showing an pvalue > 0.05 for the paired Wilcoxon Signed Rank Test. Cannot therefore assume that all the control gene-sets behave in the same general fashion. 


D) Create plots of the differences of the average log2 expression across all pairings between the SoI and controls sets in order to visualise the similarities fo the behaviour between the controls and the SoI. 
Seeing as the Friedman test came back indicating that at least one of the controls shows a different expression distribution compared to the others, cannot use the test as evidence to dismiss all controls but one. Instead we can visualise the general distribution across all controls to determine if they act in a similar fashion. 

```{r}

graph_long_data <- used_first %>% 
  dplyr::select(starts_with("Mean_diff")) %>% 
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value") 



ggplot(graph_long_data, aes(y = Value, x = Variable)) +
  geom_boxplot(varwidth = T, alpha  = 0.25, outlier.colour = "red")+
    coord_cartesian(ylim = c(0, 4)) +
  scale_y_continuous() +
  labs(x = "Unique SoI / Control Pairings ", y = "Δ Log2 Expression",
       title = "Distribution of the Set-of-Interest to Control gene-set \n Log2 Expression Difference Across All Pairings") +
  theme(
    plot.title = element_text(hjust = 0.6, size = 15),
    plot.title.position = "plot",
    axis.title = element_text(size = 13, color = "black", face = "bold"),
    axis.text = element_text(color = "black", size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.text = element_blank(),
    legend.position = "none",
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.background = element_rect(fill = "#EDEDED")
)
``` 
Interpretation: Despite that fact that the Friedman Test and the Contrast Analysis resulted in significant results (indicating a significant difference in expression difference amongst the control gene-set to SoI pairings), the plot above clearly demonstrates that the expression difference pattern between every control gene-set and the Set-of-Interest is bound in a consistent range centered between log2(2) and log2(3). And that the difference between the general expression of both sets is consistantly positive, demonstrating increased expression within the SoI compared to all control gene-sets.


E) Calculate the significance of the difference between the mean expression of the control-sets and the Set-of-Interest using a two-sided Paired Permutation Test for each control/SoI pair followed by a correction for multiple testing. 

```{r}
# Loop over each of the control datasets in the folder.
# For each control-set perform a two-sided paired permutation test with the SoI.
# Save the pvalue for each test in a dataframe.

paired_tests_list <- list()
for (i in 1:num_datasets) {
  ctrl <- paste("Mean_cntrl", i, sep ="")
  permutation_res <- perm.test(used_first$Mean_set, used_first[,get(ctrl)], alternative = "two.sided", mu = 0,
      paired = TRUE, all.perms = TRUE, num.sim = 10000, plot = FALSE, stat = mean)
  pair <- paste("Pair", i, sep ="_")
  pair_frame = data.frame(Pairing = pair, Pvalue = as.numeric(permutation_res$p.value), Corrected_Pval = as.numeric(p.adjust(permutation_res$p.value, method = "BH", n = num_datasets)))
  paired_tests_list[[i]] <- pair_frame
}

# Correct the P-value due to multiple testing by using the BH method.

full_paired <- do.call(rbind, paired_tests_list) %>% arrange(desc(Pvalue))
print(head(full_paired))

```
Result: The results show a p-value of 0 for all pairings.

Interpretation: the p-value generated by the two-sided paired permutation test demonstrates a significant difference between the means of the SoI and the Control_set for each of the control_sets. Indicating that the general expression of the genes of interest is reliably and significantly different than the general expression from the rest of the genome. 


F) Demonstrate difference by showing the individual expression distributions of the SoI
  and some (5) of the control gene-sets

```{r}
# # Select 5 control gene-sets at random to compare their distributions with that of the SoI.
# 
# controls_names <- "Mean_set"
# controls_names <- append(controls_names, unlist(lapply(controls, function(i) paste("Mean_cntrl", i, sep =""))))
# 
# # Pivot the data into a form that can be used by boxplot.
# 
# long_data <- used_first %>% 
#   pivot_longer(cols = all_of(controls_names),
#                names_to = "Variable",
#                values_to = "Value")

controls_names <- "Mean_set"
mean_cntrl_names <- unlist(lapply(controls, function(i) paste("Mean_cntrl", i, sep ="")))
controls_names <- append(controls_names, mean_cntrl_names)


# Pivot the data into a form that can be used by boxplot.

long_data <- used_first %>% 
  select(all_of(controls_names)) %>% 
  pivot_longer( cols = everything(), names_to = "Variable", values_to = "Value")

# Extract control set numbers from the Pairing column in full_paired_wilc
full_paired$Control_Num <- as.numeric(sub("Pair_", "", full_paired$Pairing))

full_paired$Sig <- ifelse (full_paired$Corrected_Pval <= 0.001, "***",
                           ifelse (full_paired$Corrected_Pval <= 0.01, "**",
                                   ifelse (full_paired$Corrected_Pval <= 0.05, "*",
"NS" )))
    

# Create a new column in full_paired_wilc to match the Variable column in long_data
full_paired$Variable <- paste("Mean_cntrl", full_paired$Control_Num, sep="")

# Merge full_paired_wilc with long_data
merged_data <- merge(long_data, full_paired, by="Variable", all.x=TRUE)

# Reorder the Variable column in merged_data
# merged_data$Variable <- factor(merged_data$Variable, levels = desired_order)


# Define the specific comparisons
specific_comparisons <- list(
  c("Mean_set", mean_cntrl_names[1]), 
  c("Mean_set", mean_cntrl_names[2]), 
  c("Mean_set", mean_cntrl_names[3]),
  c("Mean_set", mean_cntrl_names[4]), 
  c("Mean_set", mean_cntrl_names[5])
)
    
specific_comparisons_ordered <-  specific_comparisons[order(sapply(specific_comparisons, function(x) x[2]))]    

# Extract corresponding p-values for these comparisons
specific_p_values <- sapply(specific_comparisons_ordered, function(comp) {
  # Filter for rows where the Variable is either the Mean_set or the specific control
  subset_data <- merged_data[merged_data$Variable %in% comp[2],]

  # If there are multiple rows (as there should be for each pair), we take the unique p-value
  unique_pval <- unique(subset_data$Sig)

  # Return the unique p-value, or NA if not found
  if (length(unique_pval) == 1) {
    return(unique_pval)
  } else {
    return(NA)
  }
})


RNA_dif <- ggplot(long_data, aes( y = Value, x = Variable)) +
  geom_boxplot(outlier.colour = 2, fill = "gray88") +
  stat_boxplot(geom = "errorbar") +
  theme_minimal() +
  labs(title = "DISTRIBUTION OF THE MEAN EXPRESSION IN THE SET-OF-INTEREST VS VARIOUS CONTROL-SETS",
       y = expression("Expression" ~ (log[2])),) +
  scale_x_discrete(labels = c("Control Set 118","Control Set 14","Control Set 179","Control Set 195","Control Set 306", "Set of Interest")) +
  theme(
        legend.position = "None",
        axis.title.y = element_text(size= 20),
        plot.subtitle = element_text(hjust= 0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 17, angle = 45, vjust = 0.6),
        axis.text.y = element_text(size = 17),
        plot.title = element_text(hjust = 0.5,
                                  size = 25),
        panel.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(color = "black", linewidth = 0.25, linetype = 2),
        panel.border = element_rect(fill = "transparent",
                                    color = 1,
                                    linewidth = 2)) +
  geom_signif(comparisons = specific_comparisons_ordered,
               annotations = specific_p_values,
              step_increase = 0.09,
              map_signif_level = TRUE,
              textsize = 5.5,
              tip_length = 0
  ) 

```
Interpretation: the p-value and plots demonstrate that the expression pattern of the control gene-sets are significantly different from that of the SoI.

Impact on analyses: future test can now use a single control gene-set to test against as it has been shown that the controls do not differ significantly from one another and all share a similar difference with the SoI.








QUESTION 2



Data using only a single control gene-set with more rows
```{r}
main_control_num <- controls[2]


file_name <- paste("Data/RNA_Data/Full_RNA_metrics", main_control_num, ".csv", sep = "")
main_control <- fread(file_name)

main_data <- main_control[, c("Sample", "Mean_set", "Mean_cntrl", "TSS", "Hist_group", "TSS_Abbrvs")] %>%
  mutate(Mean_dif = Mean_set - Mean_cntrl) %>%
  mutate(Sample = factor(Sample, levels = Sample[order(TSS_Abbrvs)]))


```


A) Create a barplot of the samples vs their mean differences in the mean log2 RNA values coloured by TSS.

```{r}


RNA_TSS <- ggplot(main_data, aes(x = Sample, y = Mean_dif)) + 
  geom_col(alpha = 0.72, fill = "#00688B") +
  theme_minimal() +
  labs(y = expression("Expression Δ" ~ (log[2])),
       x = "Samples",
       title = "EXPRESSION DIFFERENCE ACROSS SAMPLES BY TSS") +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 17), 
    axis.text.y = element_text(size = 13),
    axis.title.x = element_blank(), 
    plot.title = element_text(hjust = 0.5, size = 23),
    plot.title.position = "plot",
    panel.background = element_rect(fill = "white"),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.5, linetype = 2),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(linewidth = 0.35, linetype = "solid", colour = "black"),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    strip.text = element_text(size = 12)) +
  guides(fill = guide_legend(ncol = 8, title = "Source Tissue Type", title.position = "top")) +
  facet_wrap(~TSS, ncol = 4, scales = "free_x", labeller = labeller(TSS = function(label) label))  

``` 
Interpretation: The positive ∆ expression between the SoI and the control gene-set is maintained across all cancer types


B) Create a barplot of the samples vs their differences in the mean log2 RNA values coloured by Hist

```{r}
my_palette <- paletteer_d("colorBlindness::paletteMartin")

main_data <- main_data %>%
  mutate( Sample = factor(Sample, levels = Sample[order(Hist_group)]))


ggplot(main_data, aes(x = Sample, y = Mean_dif, fill = Hist_group)) + 
  geom_col() +
  theme_minimal() +
  labs(y = "Δ Mean Log2 Expression",
       x = "Samples",
       title = "Expression Difference by Sample ")+
  scale_y_continuous(breaks = c(0, 1, 2, 3))+
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 14), 
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 14), 
    legend.position = "bottom",
    legend.text = element_text(size= 11),
    legend.title =element_text(hjust = 0.5,
                               size = 13),
    plot.title = element_text(hjust = 0.4,
                              size = 18),
    plot.title.position = "plot",
    panel.background = element_rect(fill = "white"),
    panel.grid.major.y = element_line(color = "black", size = 0.5, linetype = 2),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(linewidth = 0.35, linetype = "solid",
                             colour = "black"),
    panel.border = element_rect(fill = "transparent",
                                color = 1,
                                linewidth = 1)) +
  guides(fill = guide_legend(ncol = 5, 
                             title = "Histology Grouping",
                             title.position = "top")) +
   scale_fill_manual(values = my_palette)
  

``` 


C) Apply a two-sided paired permutation test between the SoI and the control gene-set while grouping by cancer type in the form of TSS and calculate the mean difference of each group


```{r}
# Get unique values of 'TSS' for grouping
unique_TSS <- unique(main_data$TSS_Abbrvs)

# Initialize a df to store the test results
Results_table <- data.frame(TSS = character(0), P_Value = numeric(0), Adj_P_Value = numeric(0), Significant_value= character(0), Mean_diff = numeric(0))

# Loop through each unique value of 'TSS'
for (i in 1:length(unique_TSS)) {
  current_TSS <- unique_TSS[i]
  
  # Subset the data for the current group
  subset_data <- main_data[main_data$TSS_Abbrvs == current_TSS, ]
  
  # Perform the paired permutation test for the current group
  test_result <- perm.test(subset_data$Mean_set, subset_data$Mean_cntrl,
                            alternative = "two.sided", mu = 0, paired = TRUE,
                            all.perms = TRUE, num.sim = 10000, plot = FALSE, stat = mean)

  # Correct the p_value using the Benjamini Hochberg correction method
  Corrected_Pval <- p.adjust(test_result$p.value, method = "BH", n = length(unique_TSS)) 
  Significant_val <- if(Corrected_Pval < 0.05) "TRUE" else "FALSE"
  
  # Calculate the mean Expression value per group
  mean_diff <- mean(subset_data$Mean_dif)
  
  # Store the result in  'Results_table' 
  Results_table <- rbind(Results_table, data.frame(TSS = current_TSS, P_Value = test_result$p.value, Adj_P_Value = Corrected_Pval,
                                                   Significant_value  = Significant_val, Mean_diff = mean_diff)) %>% arrange(desc(Mean_diff))
}

``` 
Results: All of the cancer types are shown to have a significant difference between the general Expression of the SoI and the control gene-set



D) Create barplot of each TSS group, based on mean Expression difference.

```{r}

ggplot(Results_table, aes(x = TSS, y = Mean_diff)) +
  geom_bar(stat = "identity", fill = "gray47") +
  theme_minimal() +
  labs(title = "Average Expression Difference in Set-of-Interest vs control gene-set", 
       y = "Mean Log2 Expression Difference")+
  theme(
        axis.title = element_text(size = 13),
        axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5,
                                  size = 16),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5,
                                        linetype = 1),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.2,
                                        linetype = 1),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = "transparent",
                                    color = 1,
                                    linewidth = 2)) 

``` 
Interpretation: Indicates that the Expression pattern of positive ∆ mean Expression between the SoI and the control gene-set is present across all cancer types and that there are some differences in these levels. 






