---
title: "TCGA COUNT TUMOUR - NORMAL DGEA"
format: html
editor: visual
Note: This script is for the anlysis of the Differential Gene expression between the cancerous and non-cancerous TCGA Expected count data using the SOI only.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

Load the packages

```{r}
library(dplyr)
library(knitr)
library(tidyverse) 
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(ggrepel)
```

Load the data

```{r}
original <- read.csv("Data/RNA_Data/TCGA_Norm/tcga_RSEM_Hugo_norm_count.csv")
order_data <- original[,order(colnames(original))]
```

Reduce to the SOI
```{r}

soi <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
soi_genes <-soi[,2]


transformed_data <- order_data[order_data$sample %in% soi_genes,]
rownames(transformed_data) <- NULL
```

Re-transform the data from the y = 2(x+1) values to the actual counts using x = 2^y^−1.

```{r}
untransform <- function(x) {
  return((2^x)-1)
}

count_data <- transformed_data %>%
  mutate(across(-1, untransform))
```

Transform the column names
```{r}
ids <- colnames(count_data)[-1]
participants <- sapply(ids, function(x) str_split(x, "\\.")[[1]][3])
condition <- sapply(ids, function(x) str_split(x, "\\.")[[1]][4])

column_names <- c("Samples" ,  paste(participants, condition, sep = "."))
colnames(count_data) <- column_names
```

Edits to the columns 
```{r}
# Reset the row names to the Gene names
data <- count_data %>%
  column_to_rownames(var = "Samples") 

# Remove the unwanted tissue types from the samples based on the code
codes_to_use <- c("01","02","03","04","05","08","09","10","11","12","14","40")
samples_to_use <- data %>%  select(ends_with(codes_to_use))

# Map the codes by health status (cancer <= 9 < Normal)
mapped_data <- samples_to_use
new_column_names <- colnames(samples_to_use)
new_column_names <- ifelse(as.integer(substr(new_column_names, nchar(new_column_names) - 1, nchar(new_column_names))) > 9,
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "N"),
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "C"))

colnames(mapped_data) <- new_column_names

# Remove columns with 0 values throughout
cols_zeros <- which(apply(mapped_data, 2, function(x) all(x == 0)))
mapped_data <- mapped_data[, -cols_zeros]

# Remove samples that don't have a pair 
unique_pairs <- unique(participants)

selected_columns <- c()

# Loop through each unique pair
for (pair in unique_pairs) {
  # Find the columns with the pair
  columns_with_pair <- grep(paste0("^", pair, "\\."), colnames(mapped_data))
  if (length(columns_with_pair) > 1) {
    selected_columns <- c(selected_columns, colnames(mapped_data)[columns_with_pair])
  }
}

# Select the columns with pairs from the dataframe
subset_mapped <- mapped_data[, selected_columns]
# subset_mapped <- subset_mapped[,]

```


```{r}
# convert df to matrix 
 mapped_matrix <- subset_mapped %>% as.matrix()

d0 <- DGEList(mapped_matrix, group = sapply(colnames(mapped_matrix), function(x) str_split(x, "\\.")[[1]][2]))
dim(d0)

# Remove low count genes
keep<-filterByExpr(d0,group = condition)
d0<-d0[keep,,keep.lib.size=FALSE]

dim(d0)

# Calculate the normalisation factor 
d0 <- calcNormFactors(d0, method = "TMM")

# Assess the extent of the differences library size to determine the correct approach
max_lib_size_dif <- max(d0$samples$lib.size)/min(d0$samples$lib.size)
rel_lib_size_dif <- median(d0$samples$lib.size)/min(d0$samples$lib.size)
```


```{r}

participants <- sapply(colnames(mapped_matrix), function(x) str_split(x, "\\.")[[1]][1])
condition <- sapply(colnames(mapped_matrix), function(x) str_split(x, "\\.")[[1]][2])

participants <- factor(participants)
condition <- factor(condition, levels=c("C","N"))

plotMDS(d0, pch = 1, col = ifelse(condition == "C", "red", "blue"))
```

Create the model using the condition and the participants as main effects 
```{r}
mm <- model.matrix(~ participants + condition)
y <- voom(d0, mm, plot = T)

```

Fit the linear model

```{r}
fit <- lmFit(y, mm)
fit2 <- eBayes(fit)
```


Get the table of all the genes and their differential statuses 
```{r}
express_table <- topTable(fit2, coef="conditionN", sort.by = "P", n = Inf)
```


Plot the volcano of the genes
```{r}
usable_volcano <- EnhancedVolcano(express_table,
                lab = rownames(express_table),
                x = "logFC",
                y = "P.Value",
                # selectLab = c('soi'),
                # pointSize = c(ifelse(usable_tissues$Set == "soi", 4, 1)),
                # drawConnectors = TRUE,
                # widthConnectors = 0.2,
                # xlim = c(-0.95,0.95),
                # ylim = c(0,10),
                # pCutoff = 0.05,
                FCcutoff = 1)
usable_volcano

```






PART II


Redo the above but using all genes in the DGE 

```{r}

count_data2 <- transformed_data %>%
  mutate(across(-1, untransform))
```

Transform the column names
```{r}
ids <- colnames(count_data2)[-1]
participants <- sapply(ids, function(x) str_split(x, "\\.")[[1]][3])
condition <- sapply(ids, function(x) str_split(x, "\\.")[[1]][4])
tumour_type <- sapply(ids, function(x) str_split(x, "\\.")[[1]][2])

column_names2 <- c("Samples" ,  paste(tumour_type, participants, condition,  sep = "."))
colnames(count_data2) <- column_names2
```


Load in the meta data datasets
```{r}
tss_meta <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")
abbrv_meta <- read.csv("Data/Other/TCGA_meta/bcrBatchCode.tsv", sep = "\t")


meta <- left_join(tss_meta %>% 
                    select(c("TSS.Code", "Study.Name")) %>% 
                    distinct() %>% 
                    sapply(trimws) %>% 
                    as.data.frame(),
                  abbrv_meta %>% 
                    select(c("Study.Abbreviation", "Study.Name")) %>% 
                    distinct()%>% 
                    sapply(trimws) %>% 
                    as.data.frame(), 
                  by = "Study.Name")


```



Edits to the columns 
```{r}
# Reset the row names to the Gene names
data2 <- count_data2 %>%
  column_to_rownames(var = "Samples") 


# Map the codes by health status (cancer <= 9 < Normal)
mapped_data2 <- data2
new_column_names2 <- colnames(mapped_data2)
new_column_names2 <- ifelse(as.integer(substr(new_column_names2, nchar(new_column_names2) - 1, nchar(new_column_names2))) > 9,
                           paste0(substr(new_column_names2, 1, nchar(new_column_names2) - 2), "N"),
                           paste0(substr(new_column_names2, 1, nchar(new_column_names2) - 2), "C"))

prefixes <- substr(new_column_names2,1,2)
abbrvs <- meta$Study.Abbreviation[match(prefixes,meta$TSS.Code)]
column_names_prefixes <- paste0(abbrvs, substr(new_column_names2, 3, nchar(new_column_names2)))

colnames(mapped_data2) <- column_names_prefixes


# Remove columns with 0 values throughout
cols_zeros <- which(apply(mapped_data2, 2, function(x) all(x == 0)))
mapped_data2 <- mapped_data2[, -cols_zeros]


# Remove samples that don't have a pair
unique_pairs <- unique(participants)

selected_columns <- c()


# Loop through each unique pair
for (pair in unique_pairs) {
  # Find the columns with the pair
  columns_with_pair <- grep(paste0("\\.", pair, "\\."), colnames(mapped_data2))
  if (length(columns_with_pair) > 1) {
    selected_columns <- c(selected_columns, colnames(mapped_data2)[columns_with_pair])
  }
}


# Select the columns with pairs from the dataframe
subset_mapped2 <- mapped_data2[, selected_columns]
```


```{r}
# Remove the tumour types with less than 
test <- colnames(subset_mapped2) %>% sapply( function(x) str_split(x, "\\.")[[1]][1])


```


```{r}
# convert df to matrix 
 mapped_matrix2 <- subset_mapped2 %>% as.matrix()

d1 <- DGEList(mapped_matrix2, group = sapply(colnames(mapped_matrix2), function(x) str_split(x, "\\.")[[1]][3]))
d1

# Calculate the normalisation factor 
d1 <- calcNormFactors(d1, method = "TMM")

# Assess the extent of the differences library size to determine the correct approach
max_lib_size_dif <- max(d1$samples$lib.size)/min(d1$samples$lib.size)
rel_lib_size_dif <- median(d1$samples$lib.size)/min(d1$samples$lib.size)
```

```{r}
participants <- sapply(colnames(mapped_matrix2), function(x) str_split(x, "\\.")[[1]][2])
condition <- sapply(colnames(mapped_matrix2), function(x) str_split(x, "\\.")[[1]][3])
tumour_type <- sapply(colnames(mapped_matrix2), function(x) str_split(x, "\\.")[[1]][1])

participants <- factor(participants)
condition <- factor(condition, levels=c("C","N"))
tumour_type <- factor(tumour_type)
```


Plot the multidimensional plot visually identifying the tumour type

```{r}
plotMDS(d1, pch = 1, col = as.numeric(tumour_type))
```

Plot the multidimensional plot visually identifying the sample's condition

```{r}
plotMDS(d1, pch = 1, col = ifelse(condition == "C", "red", "blue"))
```

Create the model using the condition, tumour type and the participants as main effects 
```{r}
mm2 <- model.matrix(~ participants + condition + tumour_type)
y2 <- voom(d1, mm2, plot = T)

```


Fit the linear model

```{r}
fit3 <- lmFit(y2, mm2)
fit4 <- eBayes(fit3)
```


Get the table of all the genes and their differential statuses 
```{r}
express_table2 <- topTable(fit4, coef="conditionN", sort.by = "P", n = Inf)
```


Plot the volcano of the genes
```{r}
usable_volcano2 <- EnhancedVolcano(express_table2,
                lab = rownames(express_table2),
                x = "logFC",
                y = "P.Value",
                # selectLab = c('soi'),
                # pointSize = c(ifelse(usable_tissues$Set == "soi", 4, 1)),
                # drawConnectors = TRUE,
                # widthConnectors = 0.2,
                # xlim = c(-0.95,0.95),
                # ylim = c(0,10),
                # pCutoff = 0.05,
                FCcutoff = 1)
usable_volcano2

```

