---
title: "TCGA_COUNT_TUMOUR_NORMAL_DGEA_ALLGENES"
output: html_document
date: "2024-03-18"
Note: This script is for the anlysis of the Differential Gene expression between the cancerous and non-cancerous TCGA Expected count data using all genes.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

Load the packages
```{r}
library(dplyr)
library(knitr)
library(tidyverse) 
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(ggrepel)
```


REUSABLE FUNCTIONS

```{r}
extract_element <- function(strings, index) {
  # Split each string by "." and extract the third element
  element_list <- sapply(strsplit(strings, "\\."), function(x) x[index])
  return(element_list)
}
```


COLLECTION
Load the data
```{r}
# original <- read.csv("Data/RNA_Data/TCGA_Norm/tcga_RSEM_Hugo_norm_count.csv")
order_data <- original[,order(colnames(original))]

# Metadata
tss_meta <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")
 abbrv_meta <- read.csv("Data/Other/TCGA_meta/bcrBatchCode.tsv", sep = "\t")
 meta <- left_join(tss_meta %>% 
                     select(c("TSS.Code", "Study.Name")) %>% 
                     distinct() %>% 
                     sapply(trimws) %>% 
                     as.data.frame(),
                   abbrv_meta %>%
                     select(c("Study.Abbreviation", "Study.Name")) %>% 
                     distinct()%>% 
                     sapply(trimws) %>% 
                     as.data.frame(), 
                   by = "Study.Name")

# SOI genes
soi <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
soi_genes <-soi[,2]
```


CLEANING & PREPROCESSING
Re-transform the data from the y = Log2(x+1) values to the actual counts using x = 2^y^−1.
```{r}
transformed_data <- order_data
rownames(transformed_data) <- NULL


untransform <- function(x) {#Function to convert the log transformed counts back into original counts
  return((2^x)-1)
}

count_data <- transformed_data %>%
  mutate(across(-1, untransform))
```

Edit the samples (column names) into ids so as to be able to identify the participant and the cancer type
```{r}
ids <- colnames(count_data)[-1]
participants <- extract_element(ids, 3)
condition <- extract_element(ids, 4)
tissue_type <- extract_element(ids, 2)

column_names <- c("Samples" ,  paste(tissue_type, participants, condition, sep = "."))
colnames(count_data) <- column_names
```

Edit ids so as to be able to identify the condition
```{r}
# Reset the row names to the Gene names
data <- count_data %>%
  column_to_rownames(var = "Samples") 

# Remove the unwanted sample types from the samples based on the code
codes_to_use <- c("01","02","03","04","05","08","09","10","11","12","14","40")
samples_to_use <- data %>%  select(ends_with(codes_to_use))

# Map the codes by health status (cancer <= 9 < Normal)
mapped_data <- samples_to_use
new_column_names <- colnames(samples_to_use)
new_column_names <- ifelse(as.integer(substr(new_column_names, nchar(new_column_names) - 1, nchar(new_column_names))) > 9,
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "N"),
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "C"))

prefixes <- substr(new_column_names,1,2)
abbrvs <- meta$Study.Abbreviation[match(prefixes,meta$TSS.Code)]
column_names_prefixes <- paste0(abbrvs, substr(new_column_names, 3, nchar(new_column_names)))
colnames(mapped_data) <- column_names_prefixes
```

Removal of undesirable ids/samples; those that lack expression counts and those that are not part 
of a healthy/cancerous pair.
```{r}
# Remove samples (cols) with 0 values throughout
cols_zeros <- which(apply(mapped_data, 2, function(x) all(x == 0)))
if (length(cols_zeros) > 0) {
  mapped_data_complete <- mapped_data[, -cols_zeros]
} else {
  mapped_data_complete <- mapped_data
}

# Get unique participants
unique_pairs <- unique(participants)

# List of participants with healthy/cancer paired samples
selected_columns <- c()

# Loop through each participant
for (pair in unique_pairs) {
  # Find the participants with healthy/cancer sample pairings
  pairedup <-paste0(".", pair, ".")
  columns_with_pair <- grep(pairedup, colnames(mapped_data_complete))
  matching_columns <- colnames(mapped_data_complete)[columns_with_pair]
  if (length(matching_columns) > 1 && any(grepl("\\.N$", matching_columns)) && any(grepl("\\.C$",matching_columns))) {
    # Add the found participants to the list
    selected_columns <- c(selected_columns, matching_columns)
  }
}

# Filter the data on the list of participants with healthy/cancer paired samples
subset_mapped <- mapped_data_complete[, selected_columns]

```


DGE MODELLING
Define the matrix, create its DGE object & get the normalisation library factors
```{r}
# convert df to matrix 
 mapped_matrix <- subset_mapped %>% as.matrix()

d0 <- DGEList(mapped_matrix, group = extract_element(colnames(mapped_matrix), 3))
print(paste("The size of the counts dataframe has been reduced from", 
            dim(count_data)[1], "x", dim(count_data)[2]," to ", 
            dim(d0)[1], "x", dim(d0)[2]))

# Calculate the normalisation factor 
d0 <- calcNormFactors(d0, method = "TMM")
```


Create the factors for the data. IMPORTANT: The order of the condition factor will determine how the logFCs are calculated, with the first level representing the reference (denominator) which in this case should be the healthy condition.
```{r}
participants <- extract_element(colnames(mapped_matrix), 2)
condition <- extract_element(colnames(mapped_matrix), 3)
tissue_type <- extract_element(colnames(mapped_matrix), 1)

participants <- factor(participants)
tissue_type <- factor(tissue_type)
condition <- factor(condition, levels=c("N","C"))
```


Trim low count genes
```{r}
# Remove low count genes, those with a count below 10
keep<-filterByExpr(d0,group = condition, min.count = 10)
d0<-d0[keep,,keep.lib.size=FALSE]

print(paste("The size of the dataframe is now", 
            dim(d0)[1], "x", dim(d0)[2]))
```

Plot the 2 dimensional graph of the main logFC components in terms of the condition
```{r}
plotMDS(d0, pch = 1, col = ifelse(condition == "C", "red", "blue"))
```


Plot the 2 dimensional graph of the main logFC components in terms of the tissue type

```{r}
plotMDS(d0, pch = 3, col = tissue_type)
```


Create & fit a linear model to the data.
```{r}
# Set the unique participants and the condition as the main effects.
mm <- model.matrix(~ participants + condition)
```


```{r}
# Apply voom() to transfrom the data so that it is suitable for linear modelling.
y <- voom(d0, mm, plot = T)

# Fit the linear model
fit <- lmFit(y, mm)

# smooth the standard errors
fit2 <- eBayes(fit)
```

Get the table of all the genes and their fold changes and significance 
```{r}
express_table <- topTable(fit2, coef="conditionC", sort.by = "P", n = Inf)

write.csv(express_table, "DGEA_ALLGENES_ALLTISSUES_expression_table.csv", row.names=TRUE)
```


Plot the volcano plot highlighting the GOIs
```{r}

keyvals <- ifelse(
    rownames(express_table) %in% soi_genes, 'red2',
      # ifelse(express_table$log2FoldChange > 2.5, 'gold',
        'grey63')
  keyvals[is.na(keyvals)] <- 'black'
  names(keyvals)[keyvals == 'red2'] <- 'Gene of Interest'
  names(keyvals)[keyvals == 'grey63'] <- 'Other'
  
  
usable_volcano <- EnhancedVolcano(express_table,
                lab = rownames(express_table),
                x = "logFC",
                y = "P.Value",
                title = "TCGA Differential Gene Expression",
                subtitle = "Limma-voom based",
                selectLab = soi_genes,
                colCustom = keyvals,
                pointSize = c(ifelse(rownames(express_table) %in% soi_genes, 4, 1)),
                drawConnectors = TRUE,
                widthConnectors = 0.2,
                colAlpha = 0.65,
                xlim = c(-5,5),
                pCutoff = 0.01,
                FCcutoff = 1)
usable_volcano

```