---
title: "TCGA_TUMOUR_NORMAL_ANALYSIS"
output: html_document
date: "2024-02-22"
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.my-table-style td, .my-table-style th {
    color: white !important;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir =
    "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project"
)
```

Load the libraries

```{r}
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)
library(plyr)
library(coin)
library(knitr)
library(rstatix)
library(ggpubr)
library(GGally)
library(tidyverse)
library(psych)
library(ggsci)
```

Load other options

```{r}
options(width = 120, max.print = 1000)

```

Load the RNA datasets

```{r}
set.seed(123)

soi_nrm <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_Normal_mRNA_TPM_SOI.csv")
soi_can <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
meta_data <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")

```

Configure function that builds the set of interest metrics

```{r}

calculate_summary <- function(input_dataframe) {
  df_name <- deparse(substitute(input_dataframe))

  # Calculate column means
  means <- colMeans(input_dataframe[, -c(1, 2)], na.rm = TRUE)

  # Create a new dataframe to store the summary
  summary_dataframe <- data.frame(Mean = means)

  prefix <- sub("^(.*?)_.*$", "\\1", df_name)

  colnames(summary_dataframe)[colnames(summary_dataframe) == "Mean"] <-
    paste0("geom_mean_", prefix)

  return(summary_dataframe)
}
```

Build the metrics dfs for both the tumour and the normal data and merge
the dfs on the shared Participants

```{r}
soi_combo <- merge(calculate_summary(soi_nrm) %>%
                     rownames_to_column(var = "id") %>%
                     rowwise() %>%
                     mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
                     select(-c("id")),
                   calculate_summary(soi_can) %>%
                     rownames_to_column(var = "id") %>%
                     rowwise() %>%
                     mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
                     select(-c("id")),
                   by = "Participant",
                   all = FALSE,
                   suffixes = c("_nrm", "_can"))

soi_combo <- soi_combo %>%
  dplyr::rename(geom_mean_nrm_soi = geom_mean_soi_nrm,
                           geom_mean_can_soi = geom_mean_soi_can)
```

construct a table to perform a paired wilcoxon test to assess the
differences between the two conditions using the aggregate values of the
SOI

```{r}

data <- soi_combo %>%
  tidyr::gather(key = "Condition",
         value = "Geom_mean", geom_mean_nrm_soi, geom_mean_can_soi)
print(head(data, n = 15))
```

1.  Data Assessment Plotting the distribution of the difference between
    the means to assess the symmetry and whether a wilcoxon signed rank
    test can be applied to the data. Also assessing the 'Skew' metric
    for the same purpose

```{r}
soi_combo$DIF <- soi_combo$geom_mean_can_soi - soi_combo$geom_mean_nrm_soi
ggplot(soi_combo, aes(x = "", y = soi_combo$DIF)) +
  stat_boxplot(geom = "errorbar",
               width = 0.15) +
  geom_boxplot()


describe(soi_combo) 
```

Results: Fairly symmetrical boxplot and skew of 0.4131667
Interpretation: Given that the plot is quite symmetrical and that the
skew value is between 0 and 1, I can assume that the distribution is
approximately symmetric, and can therefore use a Wilcoxon Signed Rank
test.

2.  SOI cancer-normal expression differences Boxplots to visualise the
    difference between the set of interest in cancerous data and the
    non-cancerous data

```{r}

bxp <- ggpaired(data, x = "Condition", y = "Geom_mean",
                id = "Participant",
                order = c("geom_mean_nrm_soi", "geom_mean_can_soi"),
                ylab = "Geom_mean", xlab = "Condition",
                color = "Condition",
                linetype = 0,
                line.size = 0.02,
                line.color = "grey35",
                ggtheme = theme_classic(), palette = "lancet") +
  geom_line(data = subset(
    data, Participant %in% sample(unique(data$Participant), 15)
  ),
  aes(group = Participant),
  color = "purple",
  linetype = 2)
bxp
```

Interpretation: The plot shows a slight uptrend in the mean expression
of the SOI when shifting from the Normal samples to the cancerous ones.

Perform the wilcoxon signed rank test to detect a significant difference
between the set of interest in cancerous data and the non-cancerous
data.

```{r}
stat_test <- data  %>%
  wilcox_test(Geom_mean ~ Condition, paired = TRUE, alternative = "greater") %>%
  add_significance()
stat_test

observed_val <- stat_test$p
```

Results: The difference between the mean expression levels of the SOI in
the cancerous and normal conditions is significant with a pvalue of
3.73e-21 when applying the one-sided wilcoxon signed rank test

3.  Control Sets cancer-normal expression comparison

```{r}
all_combos <- soi_combo %>%
  select(-c("DIF"))

results <- list()

num_datasets <- 547
```

Perform pairwise wilcoxon signed rank test using the paired expression
levels of the different control sets across conditions
(normal/cancerous)

```{r}

for (i in 1:num_datasets) {


  can_ctrl_file <- paste(
    "Data/RNA_Data/TCGA_TPM/TCGA_Cancer_CTRL_Sets/TCGA_TPM_RNA_Control_df", i,
    ".csv", sep = ""
  )
  can_ctrl_set <- read.csv(can_ctrl_file)

  nrm_ctrl_file <- paste(
    "Data/RNA_Data/TCGA_TPM/TCGA_Normal_CTRL_Sets/TCGA_Normal_mRNA_TPM_CTRL_Set", # nolint: line_length_linter.
    i,
    ".csv", sep = ""
  )
  nrm_ctrl_set <- read.csv(nrm_ctrl_file)


  new_combo <- merge(calculate_summary(nrm_ctrl_set) %>%
      rownames_to_column(var = "id") %>%
      rowwise() %>%
      mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
      select(-c("id")),
    calculate_summary(can_ctrl_set) %>%
      rownames_to_column(var = "id") %>%
      rowwise() %>%
      mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
      select(-c("id")),
    by = "Participant",
    all = FALSE
  )
  new_combo <- new_combo %>%
    dplyr::rename(
      !!paste0("geom_mean_nrm_", i) := geom_mean_nrm,
      !!paste0("geom_mean_can_", i) := geom_mean_can
    )

  all_combos <- all_combos %>% merge(new_combo,
    by = "Participant",
    all = FALSE
  )

  new_data <- new_combo %>%
    gather(key = "Condition", value = "Geom_mean", 2, 3)

  new_stat_res <- new_data  %>%
    wilcox_test(Geom_mean ~ Condition, paired = TRUE, alternative = "less") %>%
    add_significance()

  results[[i]] <- new_stat_res
}
```

```{r}
# all_combos <- all_combos %>%
#   dplyr::rename(geom_mean_nrm_SOI = geom_mean_nrm_soi,
#          geom_mean_can_SOI = geom_mean_can_soi
#   )
```

Combine the results into a single df and, adjust the p-value for
multiple comparisons, assess how each control set compares to the set of
interest and to the significance level.

```{r}
partial_results <- do.call(rbind, results)
partial_results <- rbind(partial_results, stat_test)
full_results <- mutate(
  partial_results, Adjusted_p = p.adjust(full_results$p, method = "BH"),
  significant_value = ifelse(Adjusted_p < 0.05, "TRUE", "FALSE"),
  greater_or_equal = ifelse(Adjusted_p <= observed_val, "TRUE", "FALSE"),
  similar_level = ifelse(p.signif == "****", "TRUE", "FALSE")
)
# 
# print(head(full_results))
```

Build table to compare the assessments made above.

```{r}
contrast_agg_count_table <- full_results[,c("significant_value","greater_or_equal","similar_level")]


summary_table <- contrast_agg_count_table %>%
  dplyr::reframe(across(everything(),
                   ~ list("TRUE" = sum(. == "TRUE"), "FALSE" = sum(. == "FALSE"))))


summary_table <- as.data.frame(t(summary_table))%>%  
  mutate(Percent_TRUE = as.numeric(V1) / (as.numeric(V1) + as.numeric(V2)),
         Fract = paste(V1, "/", as.numeric(V1) + as.numeric(V2))) %>% 
  dplyr::rename("TRUE" = V1,
         "FALSE" = V2)

kable(summary_table)
```

Resutls: of thee 547 sets

```{r}
set.seed(123)
numbers_list <- seq(1, num_datasets)
controls <- sample(1:100, 5)
```

```{r}
names <- c("Participant",
  "geom_mean_nrm_soi",
  "geom_mean_can_soi",
  unlist(lapply(numbers_list, function(i) paste("geom_mean_nrm", i, sep = "_"))),
  unlist(lapply(numbers_list, function(i) paste("geom_mean_can", i, sep = "_")))
)
select_combos <- all_combos %>% dplyr::select(c(names))

pivot_combos <- select_combos %>%
  pivot_longer(!Participant, names_to = "set", values_to = "Expression") %>%
  mutate(type = if_else(grepl("can", set), "Cancerous", "Normal"),
         set_group = sub(".*_", "", set))

```

Boxplots to visualise the difference between the set of interest in
cancerous data and the non-cancerous data

```{r}

bxp2 <- ggpaired(pivot_combos, x = "type", y = "Expression",
                 id = "Participant",
                 order = c("Normal", "Cancerous"),
                 ylab = "Expression", xlab = "Condition",
                 color = "type",
                 linetype = 0,
                 line.size = 1,
                 line.color = "purple",
                 ggtheme = theme_classic(), palette = "lancet") +
  facet_wrap(~set_group, ncol = 6, scales = "free_y") +
  # theme(strip.text.x = c("new", "new2", "nw", "n43", "53","534"),
  #       labeller = labeller())+
  geom_line(data = subset(
    pivot_combos, Participant %in% sample(unique(pivot_combos$Participant), 9)
  ),
  aes(group = Participant),
  color = "purple",
  linetype = 2
  )
```

```{r}
summary_pivot <- merge(pivot_combos %>%
  select(-c("Participant", "set")) %>%
  dplyr::group_by(type, set_group) %>%
  dplyr::summarize(mean_Expression = mean(Expression)) %>%
  tidyr::pivot_wider(names_from = type, values_from = mean_Expression) %>% 
  mutate(log2FC = Cancerous - Normal),
  full_results %>%
    mutate( set_group = sub(".*_", "", group1)) %>% 
    select(c("set_group", "Adjusted_p")),
  by = "set_group",
  all = FALSE
)
```

```{r}
p2 <- ggplot(data=summary_pivot, aes(x=log2FC, y=-log10(Adjusted_p))) + 
               geom_point() + 
               theme_minimal() + 
               geom_hline(yintercept=-log10(0.05), col="red") +
  geom_point(data = subset(
    summary_pivot, set_group == "soi"),
    aes(group = set_group),
    color = "red",
    size = 3,
    shape = 1
  )+
  geom_text(data = subset(summary_pivot, set_group == "soi"), 
            label = "Set of Interest",
            hjust = 1, vjust = 2, color = "red", size = 4)

p2
```
