---
title: "TCGA RNA CONTROL ANALYSIS"
output: html_document
date: "2024-02-20"
Note: This is the script for the analysis of the non-cancerous only TCGA data. For the combined analysis see TCGA_TUMOUR_NORMAL_ANALYSIS. For the full analysis see TCGA_TPM_mRNA_FULL_ANALYSIS. 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```
Load the libraries
```{r}
library(dplyr)
library(plyr)
library(coin)
library(knitr)
library(jmuOutlier)
library(tidyverse)
library(ggpubr)
``` 
Load other options
```{r}
getOption("max.print")
```

Load the RNA datasets


```{r}
set.seed(123)

SOI <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_Normal_mRNA_TPM_SOI.csv")
Meta_data <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")

```

Configure function that builds the set of interest metrics
```{r}

calculate_summary <- function(input_dataframe) {
  df_name <- deparse(substitute(input_dataframe))
  
  # Calculate column means
  means <- colMeans(input_dataframe[, -c(1, 2)], na.rm = TRUE)
  
  # Create a new dataframe to store the summary
  Summary_Dataframe <- data.frame( Mean = means)
  
  prefix <- sub("^(.*?)_.*$", "\\1", df_name)

  colnames(Summary_Dataframe)[colnames(Summary_Dataframe) == "Mean"] <- paste0(prefix, "_geom_mean")
  
  return(Summary_Dataframe)
}
```



```{r}
SOI_sum <- calculate_summary(SOI)

```



```{r}
start_time <- Sys.time()
# Create a table that contains all the SOI-Control_set differences for each of the controls sets
Dif_table <- data.frame(id = rownames(SOI_sum))
Mean_table <- data.frame(id = rownames(SOI_sum), SOI_mean = SOI_sum$SOI_geom_mean)


# Loop over the control datasets
num_datasets <- 25

for (i in 1:num_datasets) {
 

  ctrl_file <- paste("Data/RNA_Data/TCGA_TPM/TCGA_Normal_CTRL_Sets/TCGA_Normal_mRNA_TPM_CTRL_Set", i, ".csv", sep = "")
  ctrl_set <- read.csv(ctrl_file)

  ctrl_summary <- calculate_summary(ctrl_set)

  summary <- merge(ctrl_summary %>% rownames_to_column(var = "id"),
                 SOI_sum %>% rownames_to_column(var = "id"),
                 by = "id",
                 all = FALSE,
                 suffixes = c("", i)) %>%
  mutate(mean_dif = SOI_geom_mean - ctrl_geom_mean)

  Dif_table <- merge(Dif_table, summary %>% select(id, mean_dif),
                     by = "id", all = FALSE, suffixes = c("", i)
                     )
  Mean_table <- merge(Mean_table, summary %>% select(id, ctrl_geom_mean),
                     by = "id", all = FALSE, suffixes = c("", i)
                     )
}

end_time <- Sys.time()
cat('\n',i," iterations with a total duration of: ", end_time - start_time, "seconds")
```

```{r}
Dif_table <- Dif_table %>% dplyr::rename(mean_dif1 = mean_dif)
Mean_table <- Mean_table %>% dplyr::rename(ctrl_geom_mean1 = ctrl_geom_mean)
```



Choose a random set of 5 control gene-sets that will be used in the graphs

```{r}
set.seed(123)
controls <- sample(1:num_datasets, 5)

```



Part I. Inter-genomic Expression analysis 

QUESTION 1 


A) Create violin plots of the differences of the average log2 expression to assess the type of paired test to perform on the the pairwise groups .

```{r}

dif_pairing_names <- unlist(lapply(controls, function(i) paste("mean_dif", i, sep ="")))

violin_graph_long_data <- Dif_table[dif_pairing_names] %>%
  pivot_longer(cols = dif_pairing_names,
               names_to = "Variable",
               values_to = "Value") %>%
  filter(Value != 0)

plot1 <- ggplot(violin_graph_long_data, aes(y = Value, x = Variable)) +
  geom_violin(trim = FALSE, fill = "#778899", alpha = 0.5)+
  geom_boxplot(varwidth = T, alpha  = 0.25, outlier.colour = "red")+
  labs(x = "Unique SOI to Control-set Pairings ", y = expression("Expression Difference" ~ (log[2])),
       title = "CANCEROUS: DISTRIBUTION OF THE SET-OF-INTEREST \n TO CONTROL-SET EXPRESSION DIFFERENCE ACROSS VARIOUS PAIRINGS") +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 16, color = "black"),
    axis.text = element_text(color = "black", size = 13),
    axis.text.x = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.text = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 19),
    panel.border = element_rect(fill = "transparent", color = "#9B30FF", linewidth = 2),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    plot.title.position = "plot"
  )

print(plot1)
``` 


Interpretation: 


B) Use a Friedman's test (non-parametric version of the RMA, which is a paired t-test for more than 2 groups) to test whether the different control groups differ significantly in their distributions.


```{r}
Fried_test_data <- Dif_table %>%
  gather(key = "Pairing", value = "Difference", starts_with("mean")) %>%
  mutate(Sample = as.factor(id), Pairing = as.factor(Pairing)) %>%
  select(Sample, Pairing, Difference)

friedman_result <- coin::friedman_test(Difference ~ Pairing | Sample, data = Fried_test_data)
print(friedman_result)
```
Result: the p-value for the Friedman test shows that there is at least a significant differnece between 2 of the sets.
Interpretation: So we cannot come to the conclusion that the control sets are all similar (coming from the same distribution).


C) Construct a contrast analysis to assess each of the pairings against the general trend (excluding the current set)

```{r}
Paired_wilc_table <- Dif_table %>%
  select(starts_with("mean_dif")) %>%
  mutate_all(as.numeric)

Contrast_table <- data.frame(Set = character(0), P_Value = numeric(0), Adj_P_Value = numeric(0), Significant = character(0))

# Loop over each control gene-sets
for (i in 1:num_datasets) {

  set = paste('mean_dif', i, sep = "")
  
   Global_table <- Paired_wilc_table %>%
    mutate(global = rowMeans(select(., -set), na.rm = TRUE)) %>%
    select(set, global)

   wlcx_paired_res <- wilcox.test(Global_table[[set]], Global_table$global, paired = TRUE)
   Corrected_Pval <- wlcx_paired_res$p.value * num_datasets
   Significant_value <- if(Corrected_Pval < 0.05) "TRUE" else "FALSE"

   Contrast_table <- rbind(Contrast_table, data.frame(Set = set, P_Value = wlcx_paired_res$p.value, Adj_P_Value = Corrected_Pval,
                                                      Significant = Significant_value)) %>% arrange(desc(Adj_P_Value))
}

Contrast_agg_count_table <- Contrast_table %>% group_by(Significant) %>% dplyr::summarise(count = n())
kable(Contrast_agg_count_table)
```

Result: The Contrast analysis shows that the expression pattern of the control gene-sets differs significantly compared to the average for 526/547 with a p-value < 0.05 using the paired Wilcoxon Signed Rank Test. 




D) Create plots of the differences of the average log2 expression across all pairings between the SOI and controls sets in order to visualise the similarities fo the behaviour between the controls and the SOI. 
Seeing as the Friedman test came back indicating that at least one of the controls shows a different expression distribution compared to the others, cannot use the test as evidence to dismiss all controls but one. Instead we can visualise the general distribution across all controls to determine if they act in a similar fashion. 

```{r}

graph_long_data <- Dif_table %>% 
  dplyr::select(starts_with("mean_dif")) %>% 
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value") %>%
  filter(Value > 0)



plot2 <- ggplot(graph_long_data, aes(y = Value, x = Variable)) +
  geom_boxplot(varwidth = T, alpha  = 0.25, outlier.colour = "red")+
    coord_cartesian(ylim = c(0, 11)) +
  scale_y_continuous(breaks = seq(0, 11, by = 2)) +
  labs(x = "Set-of-Interest / Control-set Pair ", y = expression("Expression Difference" ~ (log[2])),
       title = "NON-CANCEROUS: DISTRIBUTION OF THE SET-OF-INTEREST TO \n CONTROL-SET EXPRESSION DIFFERENCE ACROSS ALL PAIRINGS") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 19),
    plot.title.position = "plot",
    axis.title = element_text(size = 16, color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.text = element_blank(),
    legend.position = "none",
    panel.border = element_rect(fill = "transparent", color = "#9B30FF", linewidth = 2),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.1, linetype = 2),
    panel.background = element_rect(fill = alpha("white", 0.1))
)

print(plot2)
``` 
Interpretation: The trend of the SOI to Control set expression difference for each of the pairings shows a fairly consistent range between 6 and 10 on a log2 scale. Demonstrating the general over-expression of the SOI compared to the rest of the genome.



E) Calculate the significance of the difference between the mean expression of the control-sets and the Set-of-Interest using a two-sided Paired Permutation Test for each control/SOI pair followed by a correction for multiple testing. 

```{r}
# Loop over each of the control datasets in the folder.
# For each control-set perform a two-sided paired permutation test with the SOI.
# Save the pvalue for each test in a dataframe.

paired_tests_list <- list()
for (i in 1:num_datasets) {
  ctrl <- paste("ctrl_geom_mean", i, sep ="")
  
  # cat("Data types of SOI_mean and", ctrl, "columns:\n")
  # str(Mean_table[c("SOI_mean", ctrl)])
  
  permutation_res <- perm.test(Mean_table$SOI_mean, Mean_table[[ctrl]], alternative = "two.sided", mu = 0,
      paired = TRUE, all.perms = TRUE, num.sim = 1000, plot = FALSE, stat = mean)
  pair <- paste("Pair", i, sep ="_")
  pair_frame = data.frame(Pairing = pair, Pvalue = as.numeric(permutation_res$p.value), Corrected_Pval = as.numeric(p.adjust(permutation_res$p.value, method = "BH", n = num_datasets)))
  paired_tests_list[[i]] <- pair_frame
}

# Correct the P-value due to multiple testing by using the BH method.

full_paired <- do.call(rbind, paired_tests_list) %>% arrange(desc(Pvalue))
kable(head(full_paired))

```
Result:

Interpretation: 



F) Demonstrate difference by showing the individual expression distributions of the SOI
  and some (5) of the control gene-sets

```{r}
controls_names <- "SOI_mean"
controls_names <- append(controls_names,  unlist(lapply(controls, function(i) paste("ctrl_geom_mean", i, sep =""))))


# Pivot the data into a form that can be used by boxplot.

long_data <- Mean_table %>%
  select(all_of(controls_names)) %>%
  pivot_longer( cols = everything(), names_to = "Variable", values_to = "Value")

# Extract control set numbers from the Pairing column in full_paired_wilc
# full_paired$Control_Num <- as.numeric(sub("Pair_", "", full_paired$Pairing))
# 
# full_paired$Sig <- ifelse (full_paired$Corrected_Pval <= 0.001, "***",
#                            ifelse (full_paired$Corrected_Pval <= 0.01, "**",
#                                    ifelse (full_paired$Corrected_Pval <= 0.05, "*",
# "NS" )))

full_paired <- full_paired %>% mutate(Control_Num = as.numeric(sub("Pair_", "", full_paired$Pairing)),
                                      Sig = ifelse (full_paired$Corrected_Pval <= 0.001, "***",
                           ifelse (full_paired$Corrected_Pval <= 0.01, "**",
                                   ifelse (full_paired$Corrected_Pval <= 0.05, "*","NS"))))


# Create a new column in full_paired_wilc to match the Variable column in long_data
full_paired$Variable <- paste("ctrl_geom_mean", full_paired$Control_Num, sep="")

# Merge full_paired_wilc with long_data
merged_data <- merge(long_data, full_paired, by="Variable", all.x=TRUE)

# Reorder the Variable column in merged_data
# merged_data$Variable <- factor(merged_data$Variable, levels = desired_order)


# Define the specific comparisons
specific_comparisons <- list(
  c(controls_names[1], controls_names[-1][1]),
  c(controls_names[1], controls_names[-1][2]),
  c(controls_names[1], controls_names[-1][3]),
  c(controls_names[1], controls_names[-1][4]),
  c(controls_names[1], controls_names[-1][5])
)

specific_comparisons_ordered <-  specific_comparisons[order(sapply(specific_comparisons, function(x) x[2]))]

# Extract corresponding p-values for these comparisons
specific_p_values <- sapply(specific_comparisons_ordered, function(comp) {
  
  # Filter for rows where the Variable is either the SOI mean or the specific control
  subset_data <- merged_data[merged_data$Variable %in% comp[2],]

  # If there are multiple rows (as there should be for each pair), we take the unique p-value
  unique_pval <- unique(subset_data$Sig)

  # Return the unique p-value, or NA if not found
  if (length(unique_pval) == 1) {
    return(unique_pval)
  } else {
    return(NA)
  }
})
```

```{r}
RNA_dif <- ggplot(long_data, aes( y = Value, x = Variable)) +
  geom_boxplot(outlier.colour = 2, fill = "#778899") +
  stat_boxplot(geom = "errorbar") +
  coord_cartesian(ylim = c(-8, 8)) +
  theme_minimal() +
  labs(title = "NON-CANCEROUS: DISTRIBUTION OF THE MEAN EXPRESSION \n IN THE SET-OF-INTEREST VS VARIOUS CONTROL-SETS",
       y = expression("Expression" ~ (log[2])),) +
  scale_x_discrete(labels = c("Control Set 118","Control Set 14","Control Set 179","Control Set 195","Control Set 306", "Set of Interest")) +
  theme(
        legend.position = "None",
        axis.title.y = element_text(size= 19),
        plot.subtitle = element_text(hjust= 0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 17, angle = 0, vjust = 0.6),
        axis.text.y = element_text(size = 17),
        plot.title = element_text(hjust = 0.5,
                                  size = 25),
        panel.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(color = "black", linewidth = 0.25, linetype = 2),
        panel.grid.minor.y = element_line(color = "black", linewidth = 0.25, linetype = 2),
        panel.border = element_rect(fill = "transparent",
                                    color = "#9B30FF",
                                    linewidth = 2)) +
  geom_signif(comparisons = specific_comparisons_ordered,
               annotations = specific_p_values,
              step_increase = 0.04,
              map_signif_level = TRUE,
              textsize = 4.5,
              tip_length = 0
  )

print(RNA_dif)

```
Interpretation:


QUESTION 2

Data using only a single control gene-set with more rows
```{r}
main_ctrl_num <- controls[2]

file_name <- paste("Data/RNA_Data/TCGA_TPM/TCGA_Normal_CTRL_Sets/TCGA_Normal_mRNA_TPM_CTRL_Set", main_ctrl_num, ".csv", sep = "")
ctrl <- read.csv(file_name)

ctrl_sum1 <- calculate_summary(ctrl)

duo_table <- merge(SOI_sum, ctrl_sum1 %>% select( ctrl_geom_mean), by = 0, all = FALSE) %>% 
  mutate(mean_dif = SOI_geom_mean - ctrl_geom_mean,
         TSS.Code = sapply(str_split(Row.names, "\\."), `[`, 2)) %>%  
  dplyr::rename(SAMPID = Row.names)


duo_table2 <- merge(duo_table, Meta_data[c("TSS.Code","Study.Name")],
                    by = "TSS.Code", all = FALSE)


```


A) Create a barplot of the samples vs their mean differences in the mean log2 RNA values segmented by Tissue Type.

```{r}

RNA_TSS <- ggplot(duo_table2, aes(x = factor(SAMPID), y = mean_dif)) + 
  geom_col(alpha = 0.72, fill = "#9B30FF") +
  theme_minimal() +
  labs(y = expression("Expression Difference" ~ (log[2])),
       x = "Samples",
       title = "NON-CANCEROUS: EXPRESSION DIFFERENCE ACROSS SAMPLES BY TUMOUR SOURCE TYPE") +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 17), 
    axis.text.y = element_text(size = 13),
    axis.title.x = element_blank(), 
    plot.title = element_text(hjust = 0.5, size = 20),
    plot.title.position = "plot",
    panel.background = element_rect(fill = "white"),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.5, linetype = 2),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line = element_line(linewidth = 0.35, linetype = "solid", colour = "black"),
    panel.border = element_rect(fill = "transparent", color = 1, linewidth = 1),
    strip.text = element_text(size = 12)) +
  guides(fill = guide_legend(ncol = 8, title = "Source Tissue Type", title.position = "top")) +
  facet_wrap(~Study.Name, ncol = 4, scales = "free_x", labeller = labeller(Study = function(label) label))  

print(RNA_TSS)
``` 

Interpretation: The positive ∆ expression between the SOI and the control gene-set is maintained across all cancer types


B) Apply a two-sided paired permutation test between the SOI and the control gene-set while grouping by cancer type in the form of Tissue Type and calculate the mean difference of each group


```{r}
# Get unique values of 'Tissue Type' for grouping
unique_Study <- unique(duo_table2$Study.Name)

# Initialize a df to store the test results
Results_table <- data.frame(Study = character(0), P_Value = numeric(0), Adj_P_Value = numeric(0), Significant_value= character(0), mean_dif = numeric(0))

# Loop through each unique value of 'Tissue Type'
for (i in 1:length(unique_Study)) {
  current_Study <- unique_Study[i]
  
  # Subset the data for the current group
  subset_data <- duo_table2[duo_table2$Study.Name == current_Study, ]
  
  # Perform the paired permutation test for the current group
  test_result <- perm.test(subset_data$SOI_geom_mean, subset_data$ctrl_geom_mean,
                            alternative = "two.sided", mu = 0, paired = TRUE,
                            all.perms = TRUE, num.sim = 10000, plot = FALSE, stat = mean)

  # Correct the p_value using the Benjamini Hochberg correction method
  Corrected_Pval <- p.adjust(test_result$p.value, method = "BH", n = length(unique_Study)) 
  Significant_val <- if(Corrected_Pval < 0.05) "TRUE" else "FALSE"
  
  # Calculate the mean Expression value per group
  mean_dif <- mean(subset_data$mean_dif)
  
  # Store the result in  'Results_table' 
  Results_table <- rbind(Results_table, data.frame(Study = current_Study, P_Value = test_result$p.value, Adj_P_Value = Corrected_Pval,
                                                   Significant_value  = Significant_val, mean_dif = mean_dif)) %>% arrange(desc(mean_dif))
}

``` 
Results: All of the cancer types are shown to have a significant difference between the general Expression of the SOI and the control gene-set


D) Create barplot of each Tissue Type group, based on mean Expression difference.

```{r}

ggplot(Results_table, aes(x = Study, y = mean_dif)) +
  geom_bar(stat = "identity", fill = "#9B30FF") +
  theme_minimal() +
  labs(title = "Average Expression Difference in Set-of-Interest vs control gene-set", 
       y = "Mean Log2 Expression Difference")+
  theme(
        axis.title = element_text(size = 13),
        axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5,
                                  size = 16),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5,
                                        linetype = 1),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.2,
                                        linetype = 1),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = "transparent",
                                    color = 1,
                                    linewidth = 2)) 

``` 

Interpretation: Indicates that the Expression pattern of positive ∆ mean Expression between the SOI and the control gene-set is present across all cancer types and that there are some differences in these levels. 