---
title: "TCGA_TUMOUR_NORMAL_ANALYSIS"
output: html_document
date: "2024-02-22"
Note: this is the script for the analysis of the cancerous vs non-cancerous samples from the TCGA data only. Does not incorporate the individual analyses. See TCGA_TPM_mRNA_FULL_ANALYSIS for the full analysis.
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.my-table-style td, .my-table-style th {
    color: white !important;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir =
    "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project"
)
```

Load the libraries

```{r}
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)
library(plyr)
library(coin)
library(rstatix)
library(ggpubr)
library(GGally)
library(tidyverse)
library(psych)
library(ggsci)
```

Load other options

```{r}
options(width = 120, max.print = 1000)

```

Load the RNA datasets

```{r}
set.seed(123)

soi_nrm <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_Normal_mRNA_TPM_SOI.csv")
soi_can <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
meta_data <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")

```

Configure function that builds the set of interest metrics

```{r}

calculate_summary <- function(input_dataframe) {
  df_name <- deparse(substitute(input_dataframe))

  # Calculate column means
  means <- colMeans(input_dataframe[, -c(1, 2)], na.rm = TRUE)

  # Create a new dataframe to store the summary
  summary_dataframe <- data.frame(Mean = means)

  prefix <- sub("^(.*?)_.*$", "\\1", df_name)

  colnames(summary_dataframe)[colnames(summary_dataframe) == "Mean"] <-
    paste0("geom_mean_", prefix)

  return(summary_dataframe)
}
```

Build the metrics dfs for both the tumour and the normal data and merge
the dfs on the shared Participants

```{r}
soi_combo <- merge(calculate_summary(soi_nrm) %>%
                     rownames_to_column(var = "id") %>%
                     rowwise() %>%
                     mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
                     select(-c("id")),
                   calculate_summary(soi_can) %>%
                     rownames_to_column(var = "id") %>%
                     rowwise() %>%
                     mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
                     select(-c("id")),
                   by = "Participant",
                   all = FALSE,
                   suffixes = c("_nrm", "_can"))

soi_combo <- soi_combo %>%
  dplyr::rename(geom_mean_nrm_soi = geom_mean_soi_nrm,
                           geom_mean_can_soi = geom_mean_soi_can)
```

construct a table to perform a paired wilcoxon test to assess the
differences between the two conditions using the aggregate values of the
SOI

```{r}

data <- soi_combo %>%
  tidyr::gather(key = "Condition",
         value = "Geom_mean", geom_mean_nrm_soi, geom_mean_can_soi)
print(head(data, n = 15))
```

1.  Data Assessment Plotting the distribution of the difference between
    the means to assess the symmetry and whether a wilcoxon signed rank
    test can be applied to the data. Also assessing the 'Skew' metric
    for the same purpose

```{r}
soi_combo$dif <- soi_combo$geom_mean_can_soi - soi_combo$geom_mean_nrm_soi
ggplot(soi_combo, aes(x = "", y = soi_combo$dif)) +
  stat_boxplot(geom = "errorbar",
               width = 0.15) +
  geom_boxplot()


describe(soi_combo) 
```

Results: Fairly symmetrical boxplot and skew of 0.4131667
Interpretation: Given that the plot is quite symmetrical and that the
skew value is between 0 and 1, I can assume that the distribution is
approximately symmetric, and can therefore use a Wilcoxon Signed Rank
test.


2.  SOI cancer-normal expression differences Boxplots to visualise the
    difference between the set of interest in cancerous data and the
    non-cancerous data

```{r}

bxp <- ggpaired(data, x = "Condition", y = "Geom_mean",
                id = "Participant",
                order = c("geom_mean_nrm_soi", "geom_mean_can_soi"),
                ylab = "Geom_mean", xlab = "Condition",
                color = "Condition",
                linetype = 0,
                line.size = 0.02,
                line.color = "grey35",
                ggtheme = theme_classic(), palette = "lancet") +
  geom_line(data = subset(
    data, Participant %in% sample(unique(data$Participant), 15)
  ),
  aes(group = Participant),
  color = "purple",
  linetype = 2)
bxp
```

Interpretation: The plot shows a slight uptrend in the mean expression
of the SOI when shifting from the Normal samples to the cancerous ones.

Perform the wilcoxon signed rank test to detect a significant difference
between the set of interest in cancerous data and the non-cancerous
data.

```{r}
stat_test <- data  %>%
  wilcox_test(Geom_mean ~ Condition, paired = TRUE, alternative = "greater") %>%
  add_significance()
stat_test

observed_val <- stat_test$p
```

Results: The difference between the mean expression levels of the SOI in
the cancerous and normal conditions is significant with a p_value of
3.73e-21 when applying the one-sided wilcoxon signed rank test. 

Interpretation: The significance test shows that the difference between the conditions is significant but whether or not it is meaningful partly depends on what the control groups show


3.  Control Sets cancer-normal expression comparison

```{r}
all_combos <- soi_combo 

results <- list()

num_datasets <- 5
```

Perform pairwise wilcoxon signed rank test using the paired expression
levels of the different control sets across conditions
(normal/cancerous)

```{r}

for (i in 1:num_datasets) {


  can_ctrl_file <- paste(
    "Data/RNA_Data/TCGA_TPM/TCGA_Cancer_CTRL_Sets/TCGA_TPM_RNA_Control_df", i,
    ".csv", sep = ""
  )
  can_ctrl_set <- read.csv(can_ctrl_file)

  nrm_ctrl_file <- paste(
    "Data/RNA_Data/TCGA_TPM/TCGA_Normal_CTRL_Sets/TCGA_Normal_mRNA_TPM_CTRL_Set", # nolint: line_length_linter.
    i,
    ".csv", sep = ""
  )
  nrm_ctrl_set <- read.csv(nrm_ctrl_file)


  new_combo <- merge(calculate_summary(nrm_ctrl_set) %>%
      rownames_to_column(var = "id") %>%
      rowwise() %>%
      mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
      select(-c("id")),
    calculate_summary(can_ctrl_set) %>%
      rownames_to_column(var = "id") %>%
      rowwise() %>%
      mutate(Participant = strsplit(id, "\\.")[[1]][3]) %>%
      select(-c("id")),
    by = "Participant",
    all = FALSE
  )
  new_combo <- new_combo %>%
    dplyr::mutate( dif = geom_mean_can - geom_mean_nrm) %>% 
    dplyr::rename(
      !!paste0("geom_mean_nrm_", i) := geom_mean_nrm,
      !!paste0("geom_mean_can_", i) := geom_mean_can,
      !!paste0("dif_", i) := dif
    )

  all_combos <- all_combos %>% merge(new_combo,
    by = "Participant",
    all = FALSE
  )

  new_data <- new_combo %>%
    gather(key = "Condition", value = "Geom_mean", 2, 3)

  new_stat_res <- new_data  %>%
    wilcox_test(Geom_mean ~ Condition, paired = TRUE, alternative = "less") %>%
    add_significance()

  results[[i]] <- new_stat_res
}

all_combos <- all_combos %>%
  dplyr::rename(geom_mean_nrm_soi = geom_mean_nrm_soi,
         geom_mean_can_soi = geom_mean_can_soi,
         dif_soi = dif
  )
```



Combine the results into a single df and, adjust the p-value for
multiple comparisons, assess how each control set compares to the set of
interest and to the significance level.

```{r}
partial_results <- do.call(rbind, results)
partial_results <- rbind(partial_results, stat_test)
full_results <- mutate(
  partial_results, Adjusted_p = p.adjust(partial_results$p, method = "BH"),
  significant_value = ifelse(Adjusted_p < 0.05, "TRUE", "FALSE"),
  greater_or_equal = ifelse(Adjusted_p <= observed_val, "TRUE", "FALSE"),
  similar_level = ifelse(p.signif == "****", "TRUE", "FALSE")
)
# 
kable(head(full_results))
```

Build table to compare the assessments made above.

```{r}
contrast_agg_count_table <- full_results[,c("significant_value","greater_or_equal","similar_level")]


summary_table <- contrast_agg_count_table %>%
  dplyr::reframe(across(everything(),
                   ~ list("TRUE" = sum(. == "TRUE"), "FALSE" = sum(. == "FALSE"))))


summary_table <- as.data.frame(t(summary_table))%>%  
  mutate(Percent_TRUE = round(as.numeric(V1) / (as.numeric(V1) + as.numeric(V2)), 3),
         Fract = paste(V1, "/", as.numeric(V1) + as.numeric(V2))) %>% 
  dplyr::rename("TRUE" = V1,
         "FALSE" = V2)

kable(summary_table)
```

Results: Out of the 547 sets :
- 110 (0.201) are also significant
- 31 (0.057) are of similar significance level (four stars)
- 0 are of equal or greater significance

Interpretation: The difference between the condition for the set of interest is quite significant when compared to that of the control sets. 

Need to get info about the directionality of the differences across the sets.



```{r}
set.seed(123)
numbers_list <- seq(1, num_datasets)
controls <- sample(1:num_datasets, 5)
```

```{r}
names <- c("Participant",
  "geom_mean_nrm_soi",
  "geom_mean_can_soi",
  unlist(lapply(controls, function(i) paste("geom_mean_nrm", i, sep = "_"))),
  unlist(lapply(controls, function(i) paste("geom_mean_can", i, sep = "_")))
)
select_combos <- all_combos %>% dplyr::select(c(names))

box_pivot_combos <- select_combos %>%
  pivot_longer(!Participant, names_to = "set", values_to = "Expression") %>%
  mutate(type = if_else(grepl("can", set), "Cancerous", "Normal"),
         set_group = sub(".*_", "", set))

```

Boxplots to visualise the difference between the set of interest in
cancerous data and the non-cancerous data

```{r}

bxp2 <- ggpaired(box_pivot_combos, x = "type", y = "Expression",
                 id = "Participant",
                 order = c("Normal", "Cancerous"),
                 ylab = "Expression", xlab = "Condition",
                 color = "type",
                 linetype = 0,
                 line.size = 1,
                 line.color = "purple",
                 ggtheme = theme_classic(), palette = "lancet") +
  facet_wrap(~set_group, ncol = 6, scales = "free_y") +
  geom_line(data = subset(
    box_pivot_combos, Participant %in% sample(unique(box_pivot_combos$Participant), 9)
  ),
  aes(group = Participant),
  color = "grey50",
  linetype = 2
  )

bxp2
```
Interpretation: The paired plot for the SOI shows not just a greater uptrend in the mean expression
of the SOI when shifting from the normal samples to the cancerous ones form about 3.8 to 3.9 on a log2 scale, but also a greater mean expression overall at around 3.8 versus the random control sets which reside at levels between -4 and -6 on a log2 scale. Demonstrating that the SOI is, when taken as an aggregate, over-expressed in general, and slightly up-regulated in cancer. 


Demonstrate this with the actual log-fold changes between the cancer and normal samples across conditions.
```{r}
names2 <- c(
  "dif_soi",
  unlist(lapply(numbers_list, function(i) paste("dif", i, sep = "_")))
)
select_combos2 <- all_combos %>% dplyr::select(all_of(names2))

pivot_main <- select_combos2 %>%
  pivot_longer(cols = everything(), names_to = "group", values_to = "Log2FC") %>% 
  group_by(group) %>%
  summarise_at(vars("Log2FC"), list(mean_log2fc = mean, std_log2fc = sd))%>%
  mutate(set = sub(".*_", "", group))

```

```{r}
summary_pivot <- merge(pivot_main %>%
  select(-c("group")),
  full_results %>%
    mutate( set = sub(".*_", "", group1)) %>% 
    select(c("set", "Adjusted_p")),
  by = "set",
  all = FALSE
)
```

```{r}
p2 <- ggplot(data=summary_pivot, aes(x=mean_log2fc, y=-log10(Adjusted_p))) + 
   geom_point(aes(size = std_log2fc), color = "grey50", fill = "grey50", alpha = 0.1) +
  geom_point(color = "black") + 
  theme_minimal() + 
  geom_hline(yintercept=-log10(0.05), col="red") +
  geom_point(data = subset(
    summary_pivot, set == "soi"),
    aes(group = set),
    color = "red",
    size = 3,
    shape = 1
  )+
  theme(legend.position = "none") +
  geom_text(data = subset(summary_pivot, set == "soi"), 
            label = "Set of Interest",
            hjust = 1, vjust = 2, color = "red", size = 4)

p2
```
Interpretation: The Volcano plot cements the fact that the set of interest demonstrates a significant adn slight general up-regulation that varies greatly from the rest of the genome. 