---
title: "TCGA_COUNT_TUMOUR_NORMAL_DGEA_ALLGENES"
output: html_document
date: "2024-03-18"
Note: This script is for the anlysis of the Differential Gene expression between the cancerous and non-cancerous TCGA Expected count data using all genes.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

Load the packages

```{r}
library(dplyr)
library(knitr)
library(tidyverse) 
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(ggrepel)
```

PLOT LOCATION FOR DGE PLOTS

```{r}
loc <- "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Plots/RNA_Plots/RNA_Plts_Part_I/DGEA_plots/"
```

REUSABLE FUNCTIONS

```{r}
extract_element <- function(strings, index) {
  # Split each string by "." and extract the third element
  element_list <- sapply(strsplit(strings, "\\."), function(x) x[index])
  return(element_list)
}
```

COLLECTION Load the data

```{r}
# Using the TCGA PanCan RSEM Expected Counts
original <- read.csv("Data/RNA_Data/TCGA_Norm/tcga_gene_expected_count.csv")
order_data <- original[,order(colnames(original))]

# Metadata
tss_meta <- read.csv("Data/Other/TCGA_meta/tissueSourceSite.tsv", sep = "\t")
 abbrv_meta <- read.csv("Data/Other/TCGA_meta/bcrBatchCode.tsv", sep = "\t")
 meta <- left_join(tss_meta %>% 
                     select(c("TSS.Code", "Study.Name")) %>% 
                     distinct() %>% 
                     sapply(trimws) %>% 
                     as.data.frame(),
                   abbrv_meta %>%
                     select(c("Study.Abbreviation", "Study.Name")) %>% 
                     distinct()%>% 
                     sapply(trimws) %>% 
                     as.data.frame(), 
                   by = "Study.Name")

# SOI genes
soi <- read.csv("Data/RNA_Data/TCGA_TPM/TCGA_mRNA_TPM_SOI.csv")
soi_genes <-soi[,2]
```

CLEANING & PREPROCESSING Re-transform the data from the y = Log2(x+1) values to the actual counts using x = 2^y^−1.

```{r}
transformed_data <- order_data
rownames(transformed_data) <- NULL


untransform <- function(x) {#Function to convert the log transformed counts back into original counts
  return((2^x)-1)
}

count_data <- transformed_data %>%
  mutate_at(vars(-1), untransform)
```

Edit the samples (column names) into ids so as to be able to identify the participant and the cancer type

```{r}
ids <- colnames(count_data)[-1]
participants <- extract_element(ids, 3)
condition <- extract_element(ids, 4)
tissue_type <- extract_element(ids, 2)

column_names <- c("Samples" ,  paste(tissue_type, participants, condition, sep = "."))
colnames(count_data) <- column_names
```

Edit ids so as to be able to identify the condition

```{r}
# Reset the row names to the Gene names
data <- count_data %>%
  column_to_rownames(var = "Samples") 

# Remove the unwanted sample types from the samples based on the code
codes_to_use <- c("01","02","03","04","05","08","09","10","11","12","14","40")
samples_to_use <- data %>%  select(ends_with(codes_to_use))

# Map the codes by health status (cancer <= 9 < Normal)
mapped_data <- samples_to_use
new_column_names <- colnames(samples_to_use)
new_column_names <- ifelse(as.integer(substr(new_column_names, nchar(new_column_names) - 1, nchar(new_column_names))) > 9,
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "N"),
                           paste0(substr(new_column_names, 1, nchar(new_column_names) - 2), "C"))

prefixes <- substr(new_column_names,1,2)
abbrvs <- meta$Study.Abbreviation[match(prefixes,meta$TSS.Code)]
column_names_prefixes <- paste0(abbrvs, substr(new_column_names, 3, nchar(new_column_names)))
colnames(mapped_data) <- column_names_prefixes
```

Removal of undesirable ids/samples; those that lack expression counts and those that are not part of a healthy/cancerous pair.

```{r}
# Remove samples (cols) with 0 values throughout
cols_zeros <- which(apply(mapped_data, 2, function(x) all(x == 0)))
if (length(cols_zeros) > 0) {
  mapped_data_complete <- mapped_data[, -cols_zeros]
} else {
  mapped_data_complete <- mapped_data
}

# Get unique participants
unique_pairs <- unique(participants)

# List of participants with healthy/cancer paired samples
selected_columns <- c()

# Loop through each participant
for (pair in unique_pairs) {
  # Find the participants with healthy/cancer sample pairings
  pairedup <-paste0(".", pair, ".")
  columns_with_pair <- grep(pairedup, colnames(mapped_data_complete))
  matching_columns <- colnames(mapped_data_complete)[columns_with_pair]
  if (length(matching_columns) > 1 && any(grepl("\\.N$", matching_columns)) && any(grepl("\\.C$",matching_columns))) {
    # Add the found participants to the list
    selected_columns <- c(selected_columns, matching_columns)
  }
}

# Filter the data on the list of participants with healthy/cancer paired samples
subset_mapped <- mapped_data_complete[, selected_columns]

```

DGE MODELLING Define the matrix, create its DGE object & get the normalisation library factors

```{r}
# convert df to matrix 
 mapped_matrix <- subset_mapped %>% as.matrix()

d0 <- DGEList(mapped_matrix, group = extract_element(colnames(mapped_matrix), 3))
print(paste("The size of the counts dataframe has been reduced from", 
            dim(count_data)[1], "x", dim(count_data)[2]," to ", 
            dim(d0)[1], "x", dim(d0)[2]))

# Calculate the normalisation factor 
d0 <- calcNormFactors(d0, method = "TMM")
```

Create the factors for the data. IMPORTANT: The order of the condition factor will determine how the logFCs are calculated, with the first level representing the reference (denominator) which in this case should be the healthy condition.

```{r}
participants <- extract_element(colnames(mapped_matrix), 2)
condition <- extract_element(colnames(mapped_matrix), 3)
tissue_type <- extract_element(colnames(mapped_matrix), 1)

participants <- factor(participants)
tissue_type <- factor(tissue_type)
condition <- factor(condition, levels=c("N","C"))
```

Trim low count genes

```{r}
# Remove low count genes, those with a count below 10
keep<-filterByExpr(d0,group = condition, min.count = 10)
d0<-d0[keep,,keep.lib.size=FALSE]

print(paste("The size of the dataframe is now", 
            dim(d0)[1], "x", dim(d0)[2]))
```

Plot the 2 dimensional graph of the main logFC components in terms of the condition

```{r}
plotMDS(d0, pch = 1, col = ifelse(condition == "C", "red", "blue"))
```

Plot the 2 dimensional graph of the main logFC components in terms of the tissue type

```{r}
plotMDS(d0, pch = 3, col = tissue_type)
```

Create & fit a linear model to the data.

```{r}
# Set the unique participants and the condition as the main effects.
mm <- model.matrix(~ participants + condition)
```

```{r}
# Apply voom() to transfrom the data so that it is suitable for linear modelling.
y <- voom(d0, mm, plot = T)

# Fit the linear model
fit <- lmFit(y, mm)

# smooth the standard errors
fit2 <- eBayes(fit)
```

Get the table of all the genes and their fold changes and significance

```{r}
express_table <- topTable(fit2, coef="conditionC", sort.by = "P", n = Inf)

write.csv(express_table, "DGEA_ALLGENES_ALLTISSUES_expression_table.csv", row.names=TRUE)
```

Plot the volcano plot highlighting the GOIs

```{r}

keyvals1 <- ifelse(
    rownames(express_table) %in% soi_genes, 'red2',
        '#8B8B83')
  keyvals1[is.na(keyvals1)] <- 'black'
  names(keyvals1)[keyvals1 == 'red2'] <- 'Gene of Interest'
  names(keyvals1)[keyvals1 == '#8B8B83'] <- 'Other'
  
  
usable_volcano <- EnhancedVolcano(express_table,
                lab = rownames(express_table),
                x = "logFC",
                y = "P.Value",
                title = "TCGA Differential Gene Expression",
                subtitle = "Combined Tissue Types",
                selectLab = soi_genes,
                colCustom = keyvals1,
                pointSize = c(ifelse(rownames(express_table) %in% soi_genes, 4, 1)),
                xlim = c(min(express_table %>%
                               filter(rownames(express_table) %in% soi_genes) %>%
                               pull(logFC)),
                         max(express_table %>%
                      filter(rownames(express_table) %in% soi_genes) %>%
                      pull(logFC))),
                ylim = c(0,
                         -log10(min(express_table %>%
                               filter(rownames(express_table) %in% soi_genes) %>%
                               pull(P.Value)))),
                drawConnectors = TRUE,
                widthConnectors = 0.2,
                colAlpha = 0.65,
                pCutoff = 0.01,
                FCcutoff = 1,
                labSize = 5,
                max.overlaps = 25,
                min.segment.length = 0.1)
usable_volcano

pdf(paste0(loc,"DGEA_VOLCANO_COMPLETE.pdf"))
usable_volcano
dev.off()

```

II. PARSED BY TISSUE/CANCER TYPE

Loop over the distinct tissue/cancer types and perform the DGE

```{r}

full_expression_table <- data.frame()

distinct_tissue_types <- unique(extract_element(colnames(subset_mapped), 1))

plotlist <- list()

for (type in distinct_tissue_types) {
  type_cols <- grep(paste0("^", type), colnames(subset_mapped), value = TRUE)
  specific_data <- subset_mapped %>% select(all_of(type_cols))

  col_number <- ncol(specific_data)
  if (col_number < 2) {
    next
  }

  # convert df to matrix
  specific_matrix <- specific_data %>% as.matrix()

  specific_d <- DGEList(specific_matrix, group = extract_element(colnames(specific_matrix), 3))

  # Calculate the normalisation factor
  specific_d <- calcNormFactors(specific_d, method = "TMM")

  # Create a factor for the variables
  participants <- extract_element(colnames(specific_matrix), 2)
  condition <- extract_element(colnames(specific_matrix), 3)
  participants <- factor(participants)
  condition <- factor(condition, levels=c("N","C"))

  # Remove low count genes, those with a count below 10
  print(paste("The size of the ", type, " dataframe was", dim(specific_d)[1], "x", dim(specific_d)[2]))
  keep <- filterByExpr(specific_d, group = condition, min.count = 10)
  specific_d <- specific_d[keep,, keep.lib.size = FALSE]
  print(paste("The size of the ", type, " dataframe is now", dim(specific_d)[1], "x", dim(specific_d)[2]))

  plotMDS(specific_d, pch = 1, col = ifelse(condition == "C", "red", "blue"), top = 25)

  # Set the unique participants and the condition as the main effects.
  specific_mm <- model.matrix(~ participants + condition)

  # Apply voom() to transform the data so that it is suitable for linear modelling.
  specific_y <- voom(specific_d, specific_mm, plot = FALSE)

  # Fit the linear model
  specific_fit <- lmFit(specific_y, specific_mm)

  # smooth the standard errors
  specific_smoothed <- eBayes(specific_fit)

  specific_express_table <- topTable(specific_smoothed, coef="conditionC", sort.by = "P", n = Inf) %>% 
    select(c("logFC", "AveExpr", "P.Value", "adj.P.Val", "B"))
  specific_express_table$Tissue <- type

  # Extract row names
  row_names <- rownames(specific_express_table)

  # Add row names as a new column to the data frame and move it to the first position
  specific_express_table <- specific_express_table %>%
    mutate(Genes = row_names) %>%
    select(Genes, everything())

  # Add the current DGE table to the full table
  full_expression_table <- rbind(full_expression_table, specific_express_table)

  keyvals <- ifelse(
    specific_express_table$Genes %in% soi_genes, 'red2', '#8B8B83')
  keyvals[is.na(keyvals)] <- 'black'
  names(keyvals)[keyvals == 'red2'] <- 'Gene of Interest'
  names(keyvals)[keyvals == '#8B8B83'] <- 'Other'
  
  
  xmini <- min(specific_express_table %>%
                filter(specific_express_table$Genes %in% soi_genes) %>%
                pull(logFC))
  
  xmaxi <- max(specific_express_table %>%
                filter(specific_express_table$Genes %in% soi_genes) %>%
                pull(logFC))
  
  ymaxi <- -log10(min(specific_express_table %>%
                filter(specific_express_table$Genes %in% soi_genes) %>%
                pull(logFC)))

  specific_volcano <- EnhancedVolcano(specific_express_table,
                                       lab = specific_express_table$Genes,
                                       x = "logFC",
                                       y = "P.Value",
                                       title = "TCGA Differential Gene Expression",
                                       subtitle = paste0(type, " Subset"),
                                       selectLab = soi_genes,
                                       colCustom = keyvals,
                                       pointSize = c(ifelse(specific_express_table$Genes %in% soi_genes, 3, 1)),
                                       xlim = c(xmini-2, xmaxi+2),
                                       ylim = c(0,ymaxi + 10/ymaxi),
                                       drawConnectors = TRUE,
                                       widthConnectors = 0.2,
                                       colAlpha = 0.65,
                                       pCutoff = 0.01,
                                       FCcutoff = 1,
                                       labSize = 5,
                                       max.overlaps = 25,
                                       min.segment.length = 0.1) 

  pdf(paste0(loc, "DGEA_VOLCANO_", type, ".pdf"))
  print(specific_volcano)  # Print the plot to the PDF
  dev.off()  # Close the PDF device

  name <- paste0(type, "_volcano")
  assign(name, specific_volcano)
}

```

Interpretation:

```{r}
write.csv(full_expression_table, "DGEA_ALLGENES_BYTISSUE_XPRSS_TBL.csv", row.names=TRUE)
```

Create the volcano plots for the group of tissues that are of most interest

```{r}
  
usable_tissues <- c("BRCA", "COAD", "HNSC", "KICH", "KIRC", "KIRP","LIHC", "LUAD", "LUSC", "PRAD", "STAD", "THCA")

usable_tissues_expression <- full_expression_table %>% filter(Tissue %in% usable_tissues)

keyvals2 <- ifelse(
  usable_tissues_expression$Genes %in% soi_genes, 'red2','#8B8B83')
  keyvals2[is.na(keyvals2)] <- 'black'
  names(keyvals2)[keyvals2 == 'red2'] <- 'Gene of Interest'
  names(keyvals2)[keyvals2 == '#8B8B83'] <- 'Other'
  
usable_tissues_expression_volcano <- EnhancedVolcano(usable_tissues_expression ,
                lab = usable_tissues_expression$Genes,
                x = "logFC",
                y = "P.Value",
                title = "TCGA Differential Gene Expression",
                subtitle = " n>10 type Samples",
                selectLab = soi_genes,
                colCustom = keyvals2,
                pointSize = c(ifelse(usable_tissues_expression$Genes %in% soi_genes,8, 3.5)),
                drawConnectors = TRUE,
                widthConnectors = 0.2,
                colAlpha = 0.65,
                pCutoff = 0.01,
                xlim = c(min(usable_tissues_expression %>%
                               filter(Genes %in% soi_genes) %>%
                               pull(logFC)),
                         max(usable_tissues_expression %>%
                      filter(Genes %in% soi_genes) %>%
                      pull(logFC))),
                ylim = c(0,
                         -log10(min(usable_tissues_expression %>%
                               filter(Genes %in% soi_genes) %>%
                               pull(P.Value)))),
                FCcutoff = 1,
                labSize = 8.5,
                max.overlaps = 25,
                min.segment.length = 0.1) +
  facet_wrap(~Tissue, scales = "free") +
  theme(strip.text = element_text(size = 15))


png(paste0(loc, "DGEA_VOLCANO_10TSS.png"), width = 3040, height = 2060)
print(usable_tissues_expression_volcano)  # Print the plot to the PDF
dev.off()  # Close the PDF device


tissues_of_interest <- c("BRCA", "HNSC", "LIHC", "LUSC")

tissues_of_interest_expression <- full_expression_table %>% filter(Tissue %in% tissues_of_interest)

keyvals3 <- ifelse(
  tissues_of_interest_expression$Genes %in% soi_genes, 'red2','#8B8B83')
  keyvals3[is.na(keyvals3)] <- 'black'
  names(keyvals3)[keyvals3 == 'red2'] <- 'Gene of Interest'
  names(keyvals3)[keyvals3 == '#8B8B83'] <- 'Other'

tissues_of_interest_expression_volcano <- EnhancedVolcano(tissues_of_interest_expression ,
                lab = tissues_of_interest_expression$Genes,
                x = "logFC",
                y = "P.Value",
                title = "TCGA Differential Gene Expression",
                subtitle = " n>10 type Samples",
                selectLab = soi_genes,
                colCustom = keyvals3,
                pointSize = c(ifelse(tissues_of_interest_expression$Genes %in% soi_genes, 8, 3.5)),
                drawConnectors = TRUE,
                widthConnectors = 0.25,
                colAlpha = 0.65,
                pCutoff = 0.01,
                xlim = c(min(usable_tissues_expression %>%
                               filter(Genes %in% soi_genes) %>%
                               pull(logFC))+0.2,
                         max(usable_tissues_expression %>%
                      filter(Genes %in% soi_genes) %>%
                      pull(logFC))-0.2),
                ylim = c(0,
                         -log10(min(usable_tissues_expression %>%
                               filter(Genes %in% soi_genes) %>%
                               pull(P.Value)))+2),
                FCcutoff = 1,
                labSize = 8.5,
                max.overlaps = 25,
                min.segment.length = 0.1) +
  facet_wrap(~Tissue, scales = "free") +
  theme(strip.text = element_text(size = 15))

png(paste0(loc, "DGEA_VOLCANO_4TSS.png"), width = 3840, height = 2160)
print(tissues_of_interest_expression_volcano)  # Print the plot to the PDF
dev.off()  # Close the PDF device

```
