---
title: "Regression parameters Assessment"
output: html_document
date: "2024-06-18"
Notes: "This is a file for the Regression Parameters Assessment, where the parameters (rna sets, depth) 
of the features are assessed and the results are presented in a table and visualized."
---

```{r}
library(dplyr)
library(knitr)
library(dplyr)
library(ggplot2)
```


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir="/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

Read in the data from the individual files
```{r}
folder <- "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/Model_output/Hyperparams/Base_models/regression/data"

full_df <- data.frame()
for (file in list.files(folder, pattern = "*.csv")){
  current_file <- read.csv(paste0(folder, "/", file)) 
  if ("Learning_Rate" %in% colnames(current_file)) {
    current_file <- current_file %>% rename("Eta" = Learning_Rate)}
  if ("Depth" %in% colnames(current_file)) {
    current_file <- current_file %>% rename("Max_depth" = Depth)}
  if ("RNA_Set" %in% colnames(current_file)) {
    current_file <- current_file %>% rename("RNA_set" = RNA_Set)}
  if ("Seed" %in% colnames(current_file)) {
    full_df <- rbind(full_df, current_file)
  } else {
    current_file <- current_file %>% mutate(., Seed = NA)
    full_df <- rbind(full_df, current_file)
  }
}

# drop the duplicates
full_df <- full_df %>% 
  distinct()


```


Create df of the unique Features by best loss
```{r}
feature_order <- c("peri_1", "peri_2", "peri_3", "peri_4", "peri_5", "peri_7", "peri_8", "peri_9", "peri_10", "peri_16_1", "peri_16_2", "peri_17", "peri_20", "peri_22_3", "ai1", "loh_hrd", "lst1")

best_loss <- full_df %>% 
  mutate(Feature = factor(Feature, levels = feature_order)) %>%
  group_by(Feature) %>% 
  slice_min(Test_RMSE, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(Feature) %>% 
  as.data.frame() %>% 
  select(Feature, RNA_set, Trees, Max_depth, Eta, Gamma, Child_weight, everything())
  
# 
# # Save the best to a csv
write.csv(best_loss, "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/Model_output/Hyperparams/Base_models/regression/best_loss.csv", row.names = FALSE)
```

Save to the hyperparams folder
```{r}
final_best_loss <- best_loss %>% 
  select(Feature, RNA_set, Trees, Max_depth, Eta, Gamma, Child_weight) %>% 
  write.csv("Data/Model_input/Hyperparameters/base_reg_hyperparams.csv", row.names = FALSE)
```


Get frequency of the weights for the best loss
```{r}
weight_loss_freq <- as.data.frame(table(best_loss$Weight_loss)) %>% 
  rename(Weight_loss = Var1)
weight_gain_freq <- as.data.frame(table(best_loss$Weight_gain)) %>% 
  rename(Weight_gain = Var1)
weight_norm_freq <- as.data.frame(table(best_loss$Weight_normal)) %>% 
  rename(Weight_normal = Var1)
combo_freq <- as.data.frame(table(best_loss$combo)) %>%
  rename(combo = Var1)
```

Get the frequency of the RNA sets for the best loss
```{r}
rna_freq <- as.data.frame(table(best_loss$RNA_Set)) %>% 
  rename(RNA_Set = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(RNA_Set) %>% 
          summarise(Test_mlogloss = mean(Test_mlogloss)), 
        by = "RNA_Set")
```

Get the detph frequency for the best loss
```{r}
depth_freq <- as.data.frame(table(best_loss$Depth)) %>% 
  rename(Depth = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(Depth) %>% 
          summarise(Test_mlogloss = mean(Test_mlogloss)), 
        by = "Depth")
```

Get the Eta frequency for the best loss
```{r}
eta_freq <- as.data.frame(table(best_loss$Eta)) %>% 
  rename(Eta = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(Eta) %>% 
          summarise(Test_mlogloss = mean(Test_mlogloss)), 
        by = "Eta")
```

Get the gamma in terms of mlogloss difference
```{r}
gamma_dif_table <- best_loss %>% 
  group_by(Gamma) %>% 
  summarise(mlogloss_diff = mean(Test_mlogloss) - mean(Trained_mlogloss),
            mean_test_mlogloss = mean(Test_mlogloss))

# plot the gamma difference and mean test mlogloss
gamma_dif_table %>% 
  ggplot(aes(x = Gamma, y = mlogloss_diff)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Gamma difference in terms of mlogloss",
       x = "Gamma",
       y = "mlogloss difference") +
  theme_minimal()
```


Save the frequency tables
```{r}
```

