---
title: "Regression parameters Assessment"
output: html_document
date: "2024-06-18"
Notes: "This is a file for the Regression Parameters Assessment, where the parameters (rna sets, depth) 
of the features are assessed and the results are presented in a table and visualized."
---

```{r}
library(dplyr)
library(knitr)
library(dplyr)
library(ggplot2)
```


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir="/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

Read in the data from the individual files
```{r}
folder <- "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/Model_output/Hyperparams/Meta_models/indiv"

full_df <- data.frame()
for (file in list.files(folder, pattern = "*.csv")){
  current_file <- read.csv(paste0(folder, "/", file)) 
  if ("Learning_Rate" %in% colnames(current_file)) {
    current_file <- current_file %>% rename("Eta" = Learning_Rate)}
  if ("Seed" %in% colnames(current_file)) {
    full_df <- rbind(full_df, current_file)
  } else {
    current_file <- current_file %>% mutate(., Seed = NA)
    full_df <- rbind(full_df, current_file)
  }
}

# drop the duplicates
full_df <- full_df %>% 
  arrange(Feature, Test_result) %>%
  distinct(Feature,
           Eta,
           Gamma,
           Max_depth,
           Child_weight,
           .keep_all = TRUE)

# Save the full_df to a csv
# write.csv(full_df, "Data/Model_output/categorical/mlogloss_measures/Full_categorical_loss_metrics_df.csv", row.names = FALSE)

```


Create df of the unique Features by best loss
```{r}
reg_feature_order <- c("peri_1", "peri_2", "peri_3", "peri_4", "peri_5", "peri_7", "peri_8", "peri_9", "peri_10", "peri_16_1", "peri_16_2", "peri_17", "peri_20", "peri_22_3", "ai1", "loh_hrd", "lst1")

best_reg_loss <- full_df %>% 
  filter(Type == "Regression") %>%
  mutate(Feature = factor(Feature, levels = reg_feature_order)) %>%
  group_by(Feature) %>% 
  slice_min(Test_result, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(Feature, Test_result) %>% 
  as.data.frame() %>% 
  select(Feature, everything(), -Type)

arm_order <- c("1p", "1q", "2p", "2q", "3p", "3q", "4p", "4q", "5p", "5q", "6p", "6q", "7p", "7q", "8p", "8q", "9p", "9q", "10p", "10q", "11p", "11q", "12p", "12q", "13q", "14q", "15q", "16p", "16q", "17p", "17q", "18p", "18q", "19p", "19q", "20p", "20q", "21q", "22q")

best_cat_loss <- full_df %>% 
  filter(Type == "Categorical") %>%
  mutate(Feature = factor(Feature, levels = arm_order)) %>%
  group_by(Feature) %>% 
  slice_min(Test_result, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(Feature, Test_result) %>% 
  as.data.frame() %>% 
  select(Feature, everything(), -Type)

best_loss <- rbind(best_cat_loss, best_reg_loss)

# # Save the best to a csv
write.csv(best_loss, "Data/Model_output/Hyperparams/Meta_models/best_loss.csv", row.names = FALSE)
```

Save as the hyperparameters to use
```{r}
best_loss_final <- best_loss %>%
  select(Feature, Trees, Eta, Gamma, Max_depth, Child_weight) %>% 
  write.csv("Data/Model_input/Hyperparameters/meta_hyperparams.csv", row.names = FALSE)

```

Get frequency of the weights for the best loss
```{r}
weight_loss_freq <- as.data.frame(table(best_loss$Weight_loss)) %>% 
  rename(Weight_loss = Var1)
weight_gain_freq <- as.data.frame(table(best_loss$Weight_gain)) %>% 
  rename(Weight_gain = Var1)
weight_norm_freq <- as.data.frame(table(best_loss$Weight_normal)) %>% 
  rename(Weight_normal = Var1)
combo_freq <- as.data.frame(table(best_loss$combo)) %>%
  rename(combo = Var1)
```

Get the frequency of the RNA sets for the best loss
```{r}
rna_freq <- as.data.frame(table(best_loss$RNA_Set)) %>% 
  rename(RNA_Set = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(RNA_Set) %>% 
          summarise(Test_result = mean(Test_result)), 
        by = "RNA_Set")
```

Get the detph frequency for the best loss
```{r}
depth_freq <- as.data.frame(table(best_loss$Depth)) %>% 
  rename(Depth = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(Depth) %>% 
          summarise(Test_result = mean(Test_result)), 
        by = "Depth")
```

Get the Eta frequency for the best loss
```{r}
eta_freq <- as.data.frame(table(best_loss$Eta)) %>% 
  rename(Eta = Var1) %>% 
  merge(., 
        best_loss %>% 
          group_by(Eta) %>% 
          summarise(Test_result = mean(Test_result)), 
        by = "Eta")
```

Get the gamma in terms of mlogloss difference
```{r}
gamma_dif_table <- full_df %>% 
  group_by(Type, Feature, Gamma) %>% 
  summarise(Result_dif = mean(Test_result) - mean(Train_result),
            mean_Test_result = mean(Test_result))

# plot the gamma difference and mean test mlogloss
gamma_dif_table %>% 
  ggplot(aes(x = Gamma, y = Result_dif, colour = Feature)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Test-Train difference in terms of Gamma",
       x = "Gamma",
       y = "Eval difference") +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(~Type, scales = "free_y")
```


Save the frequency tables
```{r}
```

