---
title: "TCGA_CNV_FEATURES"
format: html
editor: visual
Notes: This script is intended to do some exploration Arm level aneuploidies in the TCGA PanCan datasets.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

## Load the libraries

```{r}
library(dplyr)
library(mltools)
library(tidyverse)
library(data.table)
library(ggpubr)
```

Re-usable functions

```{r}
extract_element <- function(strings, index) {
  # Split each string by "." and extract the third element
  element_list <- sapply(stringr::str_split(strings, "\\-"),
                         function(x) x[index])
  return(element_list)
}
```

Import data

```{r}
ori_arm_cnv <- read_tsv("Data/CIN_Features/CNV_Data/PANCAN_ArmCallsAndAneuploidyScore_092817.txt")
```

ARM-LEVEL ANEUPLOIDIES

```{r}
# Replace the NAs with 0
cvn_arm <- ori_arm_cnv %>% replace(is.na(.), 0)

# Re-organise the columns 
cnvs_arm <- cvn_arm %>%
  column_to_rownames("Sample") %>% 
  select(-"Aneuploidy Score")
  
```

Get a sense of the data distribution for class imbalance

```{r}
freq <- cnvs_arm %>% 
  as.data.frame() %>% 
  gather(key = "arm", value = "freq", -Type) %>% 
  group_by(Type, arm) %>%
  count(freq)%>%
  as.data.frame() %>%
  spread(key = freq, value = n) %>%
  replace(is.na(.), 0)

freq_percents <- freq %>%
  mutate_at(vars(-c("arm", "Type")), as.numeric) %>%  
  mutate(total = rowSums(select(., -c("arm", "Type")))) %>%  
  mutate_at(vars(-arm,-Type, -total), funs(round(. / total*100,2))) %>%
  select(-total)  

stacked_plot <- freq_percents %>%
  rename("Loss" = "-1",
         "Normal" = "0",
         "Gain" = "1") %>%
  gather(key = "Aneuploidy", value = "Percent", -c("arm", "Type"))%>%
  mutate(Aneuploidy = factor(Aneuploidy, levels = c("Loss", "Normal", "Gain"))) %>%
  ggplot(aes(x = factor(arm, levels = c("1p", "1q", "2p", "2q", "3p", "3q", "4p", "4q", "5p", "5q", "6p", "6q", "7p", "7q", "8p", "8q", "9p", "9q", "10p", "10q", "11p", "11q", "12p", "12q", "13 (13q)", "14 (14q)", "15 (15q)", "16p", "16q", "17p", "17q", "18p", "18q", "19p", "19q", "20p", "20q", "21 (21q)", "22 (22q)")),
             y = Percent,
             fill = Aneuploidy
               )) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Arm Aneuploidies",
       subtitle = "Shown by Cancer Type",
       x = "Chromosome Arm",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.border = element_rect(fill = "transparent",
                                    color = "black",
                                    linewidth = 1.5)
        ) +
   scale_fill_brewer(palette = "Set2") +
  facet_wrap(~Type, scales = "free_x", ncol = 2)

stacked_plot


ggsave("Plots/Cancer_Type_Arm_Aneuploidies_Distribution.pdf", stacked_plot, width = 15, height = 40)
# 
```

Find the weights for the classes for each of the arms (the inverse of their prevelance)

```{r}
arm_weights <- freq %>%
  mutate(total = rowSums(select(., -arm, -Type))) %>%  
  mutate_at(vars(-arm,-Type, -total), list(~1- round(. / total,2))) %>% 
  mutate(total = rowSums(select(., -arm,-total, -Type))) %>%
  mutate_at(vars(-arm,-Type, -total), list(~ round(./total, 2))) %>%
  select(-total)

```

MDS Visualisations of the relationships between the sample/tissue types and the arm aneuploidies

```{r}
dist <- dist(cnvs_arm)
dist_matrix <- as.matrix(dist)

# # Produce the MDS object.
# mds <- cmdscale(dist)

# # Convert the MDS object to a dataframe with additional features
# mds_df <- as.data.frame(mds) %>% 
#   dplyr::mutate(cancer_type = extract_element(rownames(cnvs_arm), 2),
#          sample_type = extract_element(rownames(cnvs_arm), 4),
#          synergy_type = paste(cancer_type, sample_type, sep = "-"))

# # Create the plots of the MDS.
# sample_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2",
#   color = "sample_type", # Colour based on the sample type
#   size = 1,
#   repel = TRUE
# )

# cancer_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
#   color = "cancer_type", # Colour based on the cancer type
#   size = 1,
#   repel = TRUE) + theme(legend.position = "none")


# synergy_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
#           color = "cancer_type", # Colour based on the synergy between the cancer and sample type
#           shape = "sample_type",
#           size = 1,
#           repel = TRUE) +
#   theme(legend.position = "none")
#   # coord_cartesian(xlim = c(-35, 20),
#   #                 ylim = c(-25,25))
```

The plots don't indicate to any means of separating the data via the arm aneuploidies
