---
  title: "TCGA_CNV_FEATURES"
format: html
editor: visual
Notes: This script is intended to do some exploration of the samples in the CNV Features in the TCGA PanCan datasets.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

## Load the libraries

```{r}
library(dplyr)
library(mltools)
library(Hmisc)
library(tidyverse)
library(data.table)
library(ggpubr)
```

Re-usable functions

```{r}
extract_element <- function(strings, index) {
  # Split each string by "." and extract the third element
  element_list <- sapply(stringr::str_split(strings, "\\-"),
                         function(x) x[index])
  return(element_list)
}
```

Import data

```{r}
ori_segmnt_cnv <- read_tsv("Data/CIN_Features/CNV_Data/Chrom_seg_CNV.txt")
ori_arm_cnv <- read_tsv("Data/CIN_Features/CNV_Data/PANCAN_ArmCallsAndAneuploidyScore_092817.txt")
peri_cnv <- read.csv("/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/CIN_Features/CNV_Data/TCGA_pericentro_cnv.csv") %>%
  column_to_rownames("sampleID") 
```

ARM-LEVEL ANEUPLOIDIES

```{r}
# Replace the NAs with 0
cvn_arm <- ori_arm_cnv %>% replace(is.na(.), 0)

# Re-organise the columns 
cnvs_arm <- cvn_arm %>%
  column_to_rownames("Sample") %>% 
  select(-c("Type", "Aneuploidy Score"))
  
```

Get a sense of the data distribution for class imbalance
```{r}

old <- data.frame(Var1 = c(-1,0,1))
for (i in colnames(cnvs_arm)){
  d <- as.data.frame(table(cnvs_arm[[i]])) 
  print(d)
  old <- merge(old, d, by = "Var1")
}

```





Visualisations of the relationships

```{r}
dist <- dist(cnvs_arm)
dist_matrix <- as.matrix(dist)

# Produce the MDS object.
mds <- cmdscale(dist)

# Convert the MDS object to a dataframe with additional features
mds_df <- as.data.frame(mds) %>% 
  dplyr::mutate(cancer_type = extract_element(rownames(cnvs_arm), 2),
         sample_type = extract_element(rownames(cnvs_arm), 4),
         synergy_type = paste(cancer_type, sample_type, sep = "-"))

# Create the plots of the MDS.
sample_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2",
  color = "sample_type", # Colour based on the sample type
  size = 1,
  repel = TRUE
)

cancer_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
  color = "cancer_type", # Colour based on the cancer type
  size = 1,
  repel = TRUE) + theme(legend.position = "none")


synergy_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
          color = "cancer_type", # Colour based on the synergy between the cancer and sample type
          shape = "sample_type",
          size = 1,
          repel = TRUE) +
  theme(legend.position = "none")
  # coord_cartesian(xlim = c(-35, 20),
  #                 ylim = c(-25,25))
```

The plots don't indicate to any means of separating the data via the arm aneuploidies

PERICENTROMERIC ANEUPLOIDIES

Visualisations of the relationships

```{r}
dist <- dist(peri_cnv)
dist_matrix <- as.matrix(dist)

# Produce the MDS object.
mds <- cmdscale(dist)

# Convert the MDS object to a dataframe with additional features
mds_df <- as.data.frame(mds) %>% 
  dplyr::mutate(cancer_type = extract_element(rownames(peri_cnv), 2),
         sample_type = extract_element(rownames(peri_cnv), 4),
         synergy_type = paste(cancer_type, sample_type, sep = "-"))

# Create the plots of the MDS.
sample_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2",
  color = "sample_type", # Colour based on the sample type
  size = 1,
  repel = TRUE
)

cancer_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
  color = "cancer_type", # Colour based on the cancer type
  size = 1,
  repel = TRUE) + theme(legend.position = "none")


synergy_mds <- ggpubr::ggscatter(mds_df, x = "V1", y = "V2", 
          color = "cancer_type", # Colour based on the synergy between the cancer and sample type
          shape = "sample_type",
          size = 1,
          repel = TRUE) +
  theme(legend.position = "none")
  # coord_cartesian(xlim = c(-35, 20),
  #                 ylim = c(-25,25))
```

The plots don't indicate to any means of separating the data via the arm aneuploidies
