---
title: "Pericentromeric Segmentation"
output: html_document
date: "2024-06-01"
Notes: The script processes segmented copy number variation data to delineate pericentromeric regions, amalgamating contiguous segments and assessing copy number alterations within these regions.
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project")
```

## Load the libraries

```{r}
library(dplyr)
library(rtracklayer)
library(tidyverse)
```

Import the data
```{r}
ori_segmnt_cnv <- read_tsv("Data/CIN_Features/CNV_Data/Chrom_seg_CNV.txt")
all_segs <- import("/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Raw_Data/Other/censat.bb") %>% as.data.frame()
```

Create the pericentromeric regions by combining the adjacent gbh (gamma, beta, HSat satellite) regions which can be gaped by centromeric transition (ct) regions and selecting the longest contiguous pericentromeric region as well as any additional pericentromeric ranges that are beyond a certain threshold

```{r}
# from the T2T data select the ranges for the annotations of satellites that correspond to pericentromeric sats 

cgbh_segs <- all_segs %>% 
  filter((grepl("hsat", name)) | 
           (grepl("gsat", name)) | 
           (grepl("bsat", name)) | 
           (grepl("ct", name)) | 
           (grepl("censat", name))) %>% 
  select(c("seqnames","start", "end", "width", "name")) %>% 
  arrange(seqnames, start)

# Combine the adjacent ranges
comb_cgbh <- cgbh_segs %>%
  group_by(seqnames) %>%
  arrange(start) %>%
  mutate(end_next = lead(start),
         combined = ifelse(end == end_next & !is.na(end_next), TRUE, FALSE),
         combined_group = 0) %>% 
  select(-"end_next") %>% 
  arrange(seqnames, start )


cgbh_subset <- comb_cgbh %>%
  group_by(seqnames) %>%
  mutate(first_hsat = cummax(grepl("hsat", name)),
         last_hsat = rev(cummax(rev(grepl("hsat", name))))) %>%
  filter((first_hsat == 1) & (last_hsat == 1)) %>%
  select(-c("first_hsat","last_hsat")) %>%
  ungroup()


for (i in 2:nrow(cgbh_subset)) {
  cgbh_subset$combined_group[i] <- ifelse(
    cgbh_subset$combined[i - 1],
    cgbh_subset$combined_group[i - 1],
    cgbh_subset$combined_group[i - 1] + 1
  )
}


# Ensuring that the resulting combined regions start and end with HSat regions
adj_cgbh <- cgbh_subset %>%
  group_by(seqnames, combined_group) %>%
  mutate(first_hsat = cummax(grepl("hsat", name)),
         last_hsat = rev(cummax(rev(grepl("hsat", name))))) %>% 
  filter((first_hsat == 1) & (last_hsat == 1)) %>%
  select(-c("first_hsat","last_hsat")) %>%
  summarise(start_bp = dplyr::first(start),
            end_bp = last(end),
            width_bp = sum(width),
            name = paste(name, collapse = ", ")) %>%
  select(-c("combined_group"))

final_cgbh <- adj_cgbh %>% 
  group_by(seqnames) %>%
  filter((width_bp == max(width_bp)) | (width_bp >= 1000000)) %>%
  mutate(
    peri_id = case_when(
      n() > 1 ~ paste0("peri_", substring(as.character(seqnames), 4, nchar(as.character(seqnames))), "_", row_number()),
      TRUE ~ paste0("peri_", substring(as.character(seqnames), 4, nchar(as.character(seqnames))))),
      seqnames = substring(as.character(seqnames), 4, nchar(as.character(seqnames)))
    ) %>% 
  rename(seqnames = "chromosome") %>% 
  select(peri_id,chromosome, everything())
  
print(final_cgbh)

write.csv(final_cgbh,"/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/Other/pericentro_meta.csv")
```



Conclusion: the above yields 27 pericentromeric regions from 1 different chromosomes. The chromosomes that do not have any pericentromeric regions assigned to them include chromosomes 6,11, 12, 18 & 19, which seems to add up with what is seen in the genome viewer as well as what has been reported by Altmose (2022). The chromosomes which display more than a single elongated pericentromeric region include chromosomes:
- 13 with 3
- 15 with 2
- 16 with 2
- 21 with 2
- 22 with 4 
- Y with 2



V. Reconfigure the segmented copy number data in terms of the newly defined pericentromeric data
```{r}
cnv_long <- data.frame()
for (chrm in unique(ori_segmnt_cnv$chr)){
  if (!chrm %in% final_cgbh$chromosome){
    next
  } else {
    check_sub <- final_cgbh %>% 
      filter(chromosome == chrm)
    seg_sub <- ori_segmnt_cnv %>%
      filter(chr == chrm) %>%
      expand_grid(check_sub) %>%
      filter(
        (start >= start_bp & start < end_bp) |
          (end > start_bp & end <= end_bp) |
          (start <= start_bp & end >= end_bp)
      ) %>%
      mutate(
        overlap_start = pmax(start, start_bp),
        overlap_end = pmin(end, end_bp),
        overlap_length = overlap_end - overlap_start,
        check_sub_length = end_bp - start_bp,
        overlap_proportion = overlap_length / check_sub_length
      ) %>%
      group_by(peri_id, sampleID) %>% 
      summarise(
        cnv = sum(value * overlap_proportion) / sum(overlap_proportion),
        .groups = 'drop'  # This option removes the grouping structure after summarisation
      ) %>% 
      select(sampleID, peri_id, cnv)
  }
  cnv_long <- rbind(cnv_long, seg_sub)
}

cnv_reformat <- pivot_wider(cnv_long, names_from = peri_id, values_from = cnv)
cnv_reformat[is.na(cnv_reformat)] <- 0

write.csv(cnv_reformat, "/Users/Dyll/Documents/Education/VU_UVA/Internship/Epigenetics/Janssen_Group-UMCUtrecht/Main_Project/Data/CIN_Features/CNV_Data/TCGA_pericentro_cnv.csv")
```
